An antimicrobial drug recommender system using MALDI-TOF MS and dual-branch neural networks

Version 1

Abstract

Abstract

Timely and effective use of antimicrobial drugs can improve patient outcomes, as well as help safeguard against resistance development. Matrix-assisted laser desorption/ionization time-of-flight mass spectrometry (MALDI-TOF MS) is currently routinely used in clinical diagnostics for rapid species identification. Mining additional data from said spectra in the form of antimicrobial resistance (AMR) profiles is, therefore, highly promising. Such AMR profiles could serve as a drop-in solution for drastically improving treatment efficiency, effectiveness, and costs.

This study endeavours to develop the first machine learning models capable of predicting AMR profiles for the whole repertoire of species and drugs encountered in clinical microbiology. The resulting model can be interpreted as a drug recommender system for infectious diseases. We find that our dual-branch method delivers considerably higher performance compared to previous approaches. In addition, experiments show that the models can be efficiently fine-tuned to data from other clinical laboratories. MALDI-TOF-based AMR recommender systems can, hence, greatly extend the value of MALDI-TOF MS for clinical diagnostics.

All code supporting this study is distributed on PyPI and is packaged under: https://github.com/gdewael/maldi-nn

Introduction

In diagnostic laboratories, matrix-assisted laser desorption/ionization time-of-flight mass spectrometry (MALDI-TOF MS) is routinely used for microbial species identification (Hou et al., 2019). Usually, microbial samples only require an overnight culturing step before being analyzed with mass spectrometry (Van Veen et al., 2010; Cuénod et al., 2021). Consequently, the technology provides a time- and cost-efficient way to accurately identify the pathogen underlying an infection.

Due to the rapid evolution of antibiotic resistant strains, it is increasingly difficult to determine a treatment based on only species identity. It has been estimated that infections caused by antibiotic-resistant bacteria have caused the deaths of 1.27 million people in 2019, making AMR one of the leading causes of death on earth (Murray et al., 2022). Projections have estimated that this annual number could rise to 10 million by 2050 (O’Neill, 2016), highlighting the need for responsible antimicrobial drug use. In light of this, diagnostic laboratories will often perform various tests, such as dilution arrays or disc diffusion tests, to probe which drug will be effective (Khan et al., 2019). Such experiments typically require further culturing and are either costly, labor-intensive, time-intensive, or a mixture of the above (Humphries, 2022).

Given that MALDI-TOF spectra are already routinely used for identification, it is worth investigating to which extent they can contain further information regarding the resistance status of strains (Weis et al., 2020a). Mining this information from the spectra could help inform healthcare workers of candidate drugs. This may nullify the need for pheno-typical experiments, or (at least) direct the tests by narrowing down the choices. Furthermore, possessing a detailed resistance profile allows to treat with more specifically-working drugs (instead of broad-spectrum antibiotics) (Weis et al., 2022). Consequently, predicting resistance status from MALDI-TOF spectra could help towards the goals of antibiotic stewardship (Shlaes et al., 1997).

It has been described that some known resistance mechanisms are outside of the m/z range that MALDI-TOF spectrometers can accurately measure (Humphries, 2022). Still, it remains largely unknown to which extent co-evolved traits, such as subtle changes in metabolism caused by the resistance mechanism, can be detected by MALDI-TOF spectra. A number of studies have shown that some resistant strains can reliably be predicted from MALDI-TOF MS, either by identifying and detecting specific markers (e.g. peaks) or by learning patterns from data (see §2.1). To our knowledge, all of these studies have modeled AMR prediction for specific species-drug combinations. For this reason, they learn very specific markers of resistance, not guaranteed to extrapolate well to other drugs and species. As susceptibility rapidly evolves, it is practically impossible to perform such studies for all clinically-relevant species-drug combinations. As such, the value of aforementioned studies remains of exploratory nature with limited practical value. In addition, their performance remains limited owing to small sample sizes and, likely, the inability of MALDI-TOF spectra to fully discriminate between the characteristics of interest (Bai et al., 2017). The recently published DRIAMS dataset (Weis et al., 2022) contains phenotypic AMR data covering a wide range of species and drugs, allowing to study MALDI-TOF-based AMR prediction on an unprecedented scale.

We posit that the most pertinent challenge health-care workers face regarding AMR is to choose between all possible drugs given an infection, not whether one specific drug will be effective or not. For this reason, we argue that our models and evaluation metrics should be designed to optimally answer that question. In this study, a single model is proposed that can predict AMR for the whole range of pathogens and drugs encountered in clinical microbiology. The method jointly learns representations for antibiotic drugs and bacterial MALDI-TOF spectra. It can be used to recommend the most-likely drug to work for any drug-spectrum combination, irregardless of the species identity of the underlying pathogen. Consequently, the model is broadly-applicable and practical to use. Our contributions are as follows:

We formulate a dual-branch neural network recommender system for the prediction of AMR profiles. The model operates on MALDI-TOF spectra, as well as a representation of the candidate drug.

We evaluate multiple state-of-the-art techniques for representing drug identity in the model.

We perform evaluations by comparing our method to non-recommender system baselines.

We show that the model efficiently transfers to data from diagnostic laboratories it wasn’t trained on. Making the model easy to adopt for hospitals lacking the means and/or volume to collect large data.

Related Work

MALDI-TOF-based machine learning

The most canonical task for MALDI-TOF-based methods is species identification. Identification solutions are usually provided by the MS manufacturers and are built on large, proprietary, in-house databases (Van Belkum et al., 2012). It is unclear how these closed-source identification pipelines work, but it is likely that query spectra are directly compared to the in-house database in an approach akin to nearest neighbors (Dauwalder et al., 2023). While this approach works excellently for identification of most species, some strains remain problematic (Cao et al., 2018; Vrioni et al., 2018). Furthermore, by presumably focusing on the presence or absence of specific peaks, a lot of spectral information stands unused (Florio et al., 2018).

For various difficult prediction cases, such as strain typing, researchers often resort to machine learning (Hettick et al., 2006; Wang et al., 2018; De Bruyne et al., 2011). Stifled by a historical lack of large open data, machine learning research on MALDI-TOF data remains in its infancy. Most studies have narrow scopes and simple datasets (e.g. binary classification), only warranting standard preprocessing and off-the-shelf learning techniques (Yu et al., 2022; Zhang et al., 2023; Chung et al., 2023). Only a handful of examples exist of more advanced learning techniques specifically adapted to a MALDI-TOF-based task (Mortier et al., 2021; Weis et al., 2020a; Vervier et al., 2015). For a more thorough overview of MALDI-TOF-based machine learning, readers are referred to the review of Weis et al. (2020b).

Dual-branch neural networks

The idea of processing and combining two separate streams of information with two neural networks is applied in many fields of machine learning, collectively referred to as deep multi-target prediction (Waegeman et al., 2019; Iliadis et al., 2022).

In collaborative filtering, the goal is to predict the preference of a user to items (He et al., 2017). In its most elementary neural form, both users and items are represented by one-hot encodings, generating a model unable to make salient predictions for new users or items without having seen them during training. To solve this, a body of works exists on trying to communicate user- and item-identity to the model via side-information encoded in features (Zheng et al., 2017).

Dual-branch neural networks are also prevalent in language and vision. Recent advances in (multi-modal) contrastive learning of image (and text) representations often rely on two neural encoders to learn a matching score between two views of the same or discordant objects (Radford et al., 2021; Chen et al., 2020). Language retrieval systems typically compare input vectors with a database of key vectors, each derived from a neural network, using approximate nearest neighbor search techniques (Karpukhin et al., 2020). In biology, fields of research employing dualbranch neural networks include (1) drug-target interaction (Lee et al., 2019), (2) single-cell multi-omics analysis (Lance et al., 2022), and (3) transcription factor binding prediction (Yang et al., 2020), among countless others.

Most of these applications can, to varying extents, be interpreted as (collaborative filtering) recommender systems. For example, contrastive language-image models have been used to retrieve the most-semantically similar images to a piece of text (Beaumont, 2022).

Methods

Data

To train models, we use the recently published DRI-AMS database, consisting of 765 048 AMR measurements derived from 55 773 spectra across four different hospitals, spanning in total 74 different drugs1 (Weis et al., 2022). Every drug is characterized by a canonical SMILES string obtained from PubChem (Kim et al., 2023). As in the original DRIAMS publication, AMR measurements are binarized according to the EUCAST norms per drug. Specifically, intermediate or resistant values are assigned a positive label, and susceptible samples a negative one. Furthermore, spectra are identically processed as in the original publication. Briefly, the following steps are performed: (1) square-root transformation of the intensities, (2) smoothing using a Savitzky-Golay filter with half-window size of 10, (3) baseline correction using 20 iterations of the SNIP algorithm, (4) trimming to the 2000-20000 Da range, (5) intensity calibration so that the total intensity sums to 1, and (6) binning the intensities by summing all values in intervals of 3 Da. After preprocessing, every spectrum is represented as a 6000-dimensional vector.

The main experiments concern models that are trained on data from one hospital only (DRIAMS-A, University Hospital Basel). All spectra and measurements derived from the other three hospitals in DRI-AMS are left out for transfer learning experiments (see §4.3). Within DRIAMS-A, all spectra from before 2018 are allocated to the training set, and all spectra measured during 2018 are evenly split between validation and test set. This split in time reflects a realistic evaluation scenario, as models trained on historical data need to generalize to new patients possibly infected by newly-evolved strains. The final sizes of all splits are as follows: 409 395 labels across 28 331 spectra for the training set, 76 431 labels across 4 994 spectra for the validation set, and 76 133 labels across 4 999 spectra for the test set.

Model architecture

We formulate AMR prediction as a multi-target classification problem with side-information for both instances and targets, also referred to as dyadic prediction (Waegeman et al., 2019). In this context, let us denote a sample in the dataset 𝒟 by a triplet(si, d j, yi j), where yi j denotes the resistance label of a microbial spectrum si ∈ {1,…,n} w.r.t. a drug d j ∈ {1,…,m}. This dataset can be arranged in an incomplete score matrixY ∈ {0, 1} n×m. In what follows, the final architectural set-ups used to present the results are described. For details on hyperparameter tuning, readers are referred to Appendix B.

The model consists of two separate neural network embedders Es (⋅) and Ed (⋅) for processing the spectra and drugs, respectively. The resulting instance and target embeddings xi and t j are then combined into a single score by their scaled dot product (Rendle et al., 2020). The scaling factor, with ℎ the dimensionality of both embeddings, is inspired by the formulation of self-attention (Vaswani et al., 2017). It ensures the dot products to be of manageable magnitudes, even for large values of ℎ. This score can be used together with the sigmoid function and the cross-entropy loss to optimize the two-branch neural network to map a spectrum-drug pair to a resistance label (Iliadis et al., 2022). An overview of the model is visualized in Figure 1.

Architectural overview of the proposed model. AMR labels of spectrum-drug pairs can be represented in an incomplete matrix. A microbial sample that is susceptible to a drug is denoted by a negative label (orange), whereas positive labels (blue) signify an intermediate or resistant combination. Instance (spectrum) and target (drug) embeddings xi and t j are obtained from their respective input representations passed through their respective neural network branch. The two resulting embeddings are aggregated to a single score by their (scaled) dot product. The cross-entropy loss optimizes this score to be maximal or minimal for positive or negative combinations of microbial spectra and drugs, respectively. On the upper right-hand side, different metrics are visualized. Whereas micro ROC-AUC takes all prediction-label pairs together, the instance-wise and macro ROC-AUC compute their score per spectrum or drug, respectively, and then average.

The representations of the instance vectors xi are extracted from a neural network Es (⋅) operating on the processed and binned MALDI-TOF spectra si. Es (⋅) is parameterized by a multi-layer perceptron (MLP), consisting of a series of fully-connected layers. Between every two such layers, a series of operations consisting of (1) a GeLU activation (Hendrycks and Gimpel, 2016), (2) a dropout rate of 0.2 (Srivastava et al., 2014), and (3) layer normalization (Ba et al., 2016), is applied. We include multiple model sizes in our final results (Table 1). To make comparisons easier, all models output the same number of hidden dimensions that are used in the dot product, xi ∈ ℝ64.

Drug identity can be communicated to the model in a number of ways. In this work, we study the following different input representations d j and embedder Ed(⋅) combinations:

As indices in a one-hot encoding paired with a single linear layer.

As Extended Connectivity Fingerprints paired with a single linear layer.

As DeepSMILES strings (O’Boyle and Dalke, 2018) paired with a 1D convolutional neural network (CNN).

As DeepSMILES strings paired with a gated recurrent unit neural network (GRU).

As DeepSMILES strings paired with a transformer neural network.

As images paired with a 2D CNN.

As rows of a pre-computed string kernel on the SMILES strings (LINGO (Vidal et al., 2005)), paired with a single linear layer.

For all these combinations, the embedder outputs target embeddings t j ∈ ℝ64. For more details on the different drug embedders and their hyperparam-\ eters (as well as their tuning), see Appendix B. For every combination of spectrum embedder (four sizes: S, M, L, and XL) and drug embedder (seven types), four different learning rates ({1e-4, 5e-4, 1e-3, 5e-3}) are tested. 1e−10 For all these different combinations, three models are trained (using different random seeds for model initialization and batching of data). For every spectrum and drug embedder combination, only results from the best learning rate are presented; that is, the learning rate resulting in the best average validation micro ROC-AUC for that combination.

All models are trained with the Adam optimizer (Kingma and Ba, 2014) for a maximum of 50 epochs with a batch size of 128. A linear learning rate warm-up over the first 250 steps is applied, after which the rate is kept constant. As every epoch constitutes one pass over every label and, hence, multiple passes over every individual drug and spectrum, a branch can technically already be overfitting before the end of the first epoch. Because of this, performance on the validation set is checked every tenth of an epoch. Training is halted early when validation micro ROC-AUC hasn’t improved for 10 validation set checks. The checkpoint of the best performing model (in terms of validation micro ROC-AUC) is used as the final model.

Results

The following section will first relay the results of the different dual-branch model configurations. After, a comprehensive comparison is made with baselines consisting of specialist models trained on specific species-drug combinations, as is now common practice in existing works. Finally, the models’ capabilities and representations are examined through transfer learning and embeddings.

Encoding species and drugs effectively

Figure 2 shows the performance of all trained models in terms of their areas under the receiver operating characteristic curve (ROC-AUC). The ROC-AUC measures the probability that any positive (resistant or intermediate) sample is assigned a higher predicted probability of being positive as compared to any negative (susceptible) sample. The micro ROC-AUC computes this probability for any pair of positive and negative samples, whereas the instance-wise ROC-AUC only compares pairs within the same spectrum, respectively2 (Waegeman and Dembczynski, 2018). In a realistic scenario, an AMR recommender system, such as this one, would be used to evaluate which drugs work for new patients. Instance-wise metrics reflect on the average performance across drugs for a single patient. The instance-wise ROC-AUC, for instance, measures the average quality of the ranking of suggested drugs to a patient.

Barplots showing test performance results for all trained models. Errorbars represent the standard deviation over three random seeds. The x-axis and colors show the different drug and spectrum embedders, respectively.

From Figure 2 it can be seen that, in general, performance differences between model configurations occupy a small margin. However, trends can still be found. One-hot drug embedders typically outperform those using molecular structure information. We offer two hypotheses for this observation: (1) the number of drugs in the training dataset is too small to learn salient representations based on molecular information, and/or (2) wildly different resistance profiles may be obtained from very similar molecular structures. In the latter case, including molecular structure in the model may actually hinder the model, rather than helping it. Among the other drug embedders, the best performance is found in models using Morgan fingerprints. On the spectrum embedder side, it is observed that the small or medium-sized variants typically perform best.

Performance in terms of Macro ROC-AUC and Instance-wise Prec@1 of the negative class can be found in Appendix C Figure 8. The Macro ROC-AUC averages the ROC-AUC for every individual drug. The Prec@1 evaluates how often the top-ranked prediction is correct. The Prec@1 of the negative class, hence, reports the proportion of cases for which the “most-likely susceptible drug” prediction is actually an effective one. In a scenario where the top recommended drug is always administered, it corresponds to the percentage of correctly-suggested treatments. It can be observed that, in terms of Macro ROC-AUC, Morgan fingerprints (and sometimes other drug embedders) outperform one-hot drug embedders. Hence, molecular structures are a useful inductive bias to learn salient representations for drugs not encountered much in the data. For the Instance-wise Prec@1, a similar observation can be made. The full list of performances can be found in Appendix C Table 4.

In Appendix C Figure 9, the performance of the spectrum embedder sizes is compared against a linear baseline. The linear baseline uses the same pre-processed input spectrum representation, but only uses a single linear combination to produce an embedding. For this comparison, only the one-hot and Morgan fingerprint drug embedders are used, as they produce the best-performing models overall. Models using non-linear multi-layer spectrum embedders obtain considerably better performance over linear embedders.

Dual-branch modeling improves AMR prediction

Previous studies have studied AMR prediction in specific species-drug combinations (Weis et al., 2020b). To the best of our knowledge, this study represents the first attempt at modeling MALDI-TOF-based AMR prediction across a wide array of species and drugs using a single model. For this reason, it is useful to compare how the dual-branch setup weighs up against training separate models for separate species and drugs. To test this, binary logistic regression, XGBoost (Chen and Guestrin, 2016) and MLPs are trained for the 200 combinations of species and drugs for which most training labels are present. The tested MLPs come in the same four sizes as the spectrum branches of the dual-branch models. Other than having an output node of size 1 for binary classification, they share all hyperparameters with the tested spectrum branches.

There exist many species-drug combinations for which there are either only positive or only negative labels. As it is impossible to train and evaluate models for these cases, models are trained only for the 200 most-occurring combinations for which both labels are present in the training, validation and test set. We refer to these models as “specialists”, as they are optimized to recognize patterns for specific cases of AMR. For details on the training and tuning procedure of all baselines, see Appendix B.2.2.

The performances of specialist models are compared to two dual-branch recommenders (Table 2). Two representative dual-branch models are chosen based on their validation micro ROC-AUC. One model is chosen among those using one-hot drug embedders, and another is chosen among those not using a one-hot drug embedder. Performance is computed on the subset of labels spanning the 200 most-common species-drug combinations. This subset spans 4017 spectra, 35 drugs, and 53503 labels (covering 70.28% of the original test set). Dual-branch recommenders outperform specialist baselines on all but one metric. Logistic regression baselines result in the best average ROC-AUC for individual species-drug combinations. By all other metrics, dual-branch recommenders outshine a collection of specialist models. These results are perhaps not surprising, as one could argue that either type of models is designed to be optimal for the metric they outperform the other in. It’s illustrated that, when the question is to choose between drugs for a patient (i.e. instance-wise metrics), a model designed as a recommender will out-perform binary classification models trained to predict AMR for specific drugs. On the other hand, when the pathogen is already known, and only one drug is in consideration, a binary classification model will be optimal.

Multi-task learning can explain why dual-branch models outperform a collection of specialists (Waegeman et al., 2019; Caruana, 1997). As species will often share some common biological processes causing AMR, representations learned on one species may strengthen those for another. In doing so, the combination of data from multiple species in one model helps to produce more expressive spectrum representations overall. Similarly, many antimicrobial drugs share common structure and mechanisms. As such, combining the data from different drugs into one dataset and model may similarly lead to performance gains.

Efficient transfer learning to new hospitals

An AMR prediction model trained using data from one hospital may not be suitable for use in other hospitals for several reasons. First, protocols such as sample preparation and culturing media differ from hospital to hospital, resulting in systematic differences in MALDI-TOF spectra (Weis et al., 2022). Second, epidemiology is spatially varied. Drug-resistant clades may be prevalent in one region or country, but absent in another (Humphries, 2022). Finally, the MALDI-TOF instruments themselves may also be specific to the hospital and influence the readout. This influences prediction models, as a hospital-specific effect is reported by the study introducing the DRIAMS dataset (Weis et al., 2022). They find that models typically perform best when trained with data from the same hospital. Here, hospital transferability is studied in the context of transfer learning.

Data from DRIAMS-B, -C and -D, are split into training, validation and test set. The train set for these hospitals consists of 1000 randomly-drawn spectra, simulating a small data scenario where the hospital has not spent considerable efforts in data collection. The remaining spectra for all three hospitals are evenly split among validation and test set.

For all three hospitals, we train models in the same way as previously (see §3.2). A comparison is made between fine-tuning starting from models trained on DRIAMS-A (i.e. models from previous sections) and dual-branch models trained from scratch (Figure 3). Over all three hospitals, models fine-tuned from a DRIAMS-A checkpoint generally outperform models trained from scratch. This trend holds true over different numbers of spectra available in the fine-tuning set. In general, it can be seen that pre-trained models require very little fine-tuning spectra to obtain performances in the same order of magnitude as with DRIAMS-A (§4.1). Curiously, whereas the one-hot drug embedders typically outperform Morgan fingerprints-based models on DRIAMS-A, here, we see an opposite trend. This effect may be explained by the fact that not all hospitals use the same set of drugs. Whereas a Morgan fingerprint-based drug embedder may (to an extent) instantly generalize to new drugs, a one-hot drug embedder needs to learn a representation from scratch for every drug it hasn’t seen before. Performance comparisons of the same models in terms of other metrics are shown in Appendix C Figure 10

Transfer learning of DRIAMS-A models to other hospitals. Errorbands show the standard deviation over three runs. Results in terms of other evaluation metrics are shown in Appendix C Figure 10.

Lowering the amount of data required is paramount to expedite the uptake of AMR models in clinical diagnostics. The transfer learning qualities of dual-branch models may be ascribed to multiple properties. First of all, since different hospitals use much of the same drugs, transferred drug embedders allow for expressively representing drugs out of the box. Secondly, owing to multi-task learning, even with a limited number of spectra, a considerable fine-tuning dataset may be obtained, as all available data is “thrown on one pile”.

MALDI-TOF spectra embeddings

To investigate what the dual-branch models have learned to represent, MALDI-TOF spectra embeddings are examined. For this purpose, the model with the best validation micro ROC-AUC is used, which is a model using a medium-size spectrum embedder and one-hot drug embedder. Here, we visualize the embeddings xi ∈ ℝ64 of all test set spectra from the 25 most-occurring pathogens. To visualize in a 2-dimensional space, UMAP is applied (using default parameters apart from min_dist=0.753). Figure 4 shows the resulting embeddings, colored by species identity, as well as by their AMR status to a selection of drugs.

UMAP scatterplots of test set MALDI-TOF spectra embeddings xi. Only embeddings belonging to the 25 most-occurring species in the test set are shown. The panels on the right show the same embeddings as on the left, but colored according to its AMR status to a certain drug. The four displayed drugs are selected based on a ranking of the product of the number of positive and negative labels . In this way, the drugs that have a lot of observed labels, both positives and negatives, are displayed.

The MALDI-TOF embeddings are grouped primarily per species. This shows that, without being instructed to discriminate between species, the model has learned to group spectra of the same species together. Furthermore, species under the same genus are typically grouped close together, illustrating that the model can pick up hierarchical relations in the tree of life from the data. Within species clusters, the AMR status subplots show that samples are often grouped according to their resistance. For example, for S. epidermidis and S. aureus, multidrug resistant variants clearly form subclusters. In addition, the cluster of E. coli spectra shows a clear tail with samples resistant to ciprofloxacin. UMAP embedding plots for other drugs are shown in Appendix C Figure 11.

Discussion

Prior work on AMR prediction has always modeled within the boundaries of one clade and drug(class), using standard machine learning practices. This work differentiates itself from others by constructing one model for the whole range of pathogens and drugs encountered in clinical diagnostics. We propose to model AMR prediction via dual-branch neural networks, producing a novel MALDI-TOF-based AMR recommender system. The proposed models come with improved performance over the approaches taken in previous works. Only when evaluated on a perpathogen-and-drug basis, it makes more sense to have separate models. It is worth noting, though, that the dual-branch models can also be fine-tuned to similar “specialist” models on a certain of subset of data (spanning, for example, one species and drug). Aside from this point, we argue that, in any case, instance-wise metrics are more appropriate for AMR prediction. They judge the quality of predictions on a patient-per-patient basis, as it would happen in usage of such models in clinical diagnostics.

We postulate that the performance of the proposed models is still limited due to (1) lacking a MALDI-TOF-specific learning architecture, (2) collection of more data, especially on rarely-encountered species and drugs, and (3) inherent technological limitations of MALDI-TOF MS. Whilst the former is the subject of further machine learning research, the latter two can be considered by equipping the model with some notion of uncertainty, epistemic and aleatoric, respectively (Hüllermeier and Waegeman, 2021). In medical decision-making applications, effective uncertainty estimates would be an invaluable tool to aid understanding the models’ predictions. A fourth factor to consider is that perfect test set performance may also be unattainable due to labeling errors. This comes as a consequence of (1) error-prone laboratory measurements of MIC values, and (2) the fact that EUCAST norms change over time, resulting in outdated label thresholds for historical data.

As bacterial strains readily adapt resistance to new and frequently-used antibiotics, it is impossible for a single machine learning model to maintain its performance over time. Due to this obvious need for continual data collection and online machine learning, it is clear that ML for AMR prediction will prove most valuable when integrated tightly in the inner workings of healthcare (Lee and Lee, 2020).

It stands to reason that blindly following the recommender system’s predictions spells misery. For example, healthcare practitioners should additionally take into account host-specific factors such as patient age, medical history, and concurrent medication. Additionally, as the model is trained on the whole repertoire of antimicrobial drugs, it will have learnt that broad-spectrum antibiotics are typically effective. Hence, it may overrecommend their use. As a consequence, the model’s proposed treatment strategies may not be aligned with antibiotic stewardship, instead exacerbating the very issue it is designed to mitigate. To tackle this problem, one could downweigh the prediction probabilities of undesirable drugs, or, alternatively, train a dual-branch model on only more-specifically-working drugs.

In summary, this study serves as the first proof-of-concept for large MALDI-TOF-based antimicrobial drug recommenders. In this context, we highlight the need for appropriate metrics, proposing that instance-wise metrics are most suitable. Extensive experiments on our proposed dual-branch model allow us to assemble some conclusions w.r.t. its use. Firstly, given the present data at hand, medium-sized MLP spectrum embedders (counting 3.2M weights) generally perform best. Second, embedding drugs without structural information (by one-hot encoding) works best for the larger DRIAMS-A dataset. For the smaller datasets used in the transfer learning experiments, the structural inductive bias lent to the model via Morgan fingerprints delivers best results. Our experiments demonstrate that dual-branch recommenders outperform non-recommender baselines on relevant metrics. In the above discussion, some considerations are listed w.r.t. its practical implementation in health-care. Taken together, this work demonstrates the potential of AMR recommenders to greatly extend the value of MALDI-TOF MS for clinical diagnostics.

eLife Assessment

This valuable study presents a machine learning model to recommend effective antimicrobial drugs from patients' samples analysed with mass spectrometry. While the proposed approach of training a single model across different bacterial species and drugs seems promising, the comparison with baselines and related work is incomplete. With the evaluation part strengthened, this paper would be of interest to computational biologists, microbiologist, and clinicians.

Reviewer #1 (Public Review):

Summary:

De Waele et al. reported a dual-branch neural network model for predicting antibiotic resistance profiles using matrix-assisted laser desorption/ionization time-of-flight (MALDI-TOF) mass spectrometry data. Neural networks were trained on the recently available DRIAMS database of MALDI-TOF mass spectrometry data and their associated antibiotic susceptibility profiles. The authors used a dual branch neural network approach to simultaneously represent information about mass spectra and antibiotics for a wide range of species and antibiotic combinations. The authors showed consistent performance of their strategy to predict antibiotic susceptibility for different spectrums and antibiotic representations (i.e., embedders). Remarkably, the authors showed how small datasets collected at one location can improve the performance of a model trained with limited data collected at a second location. Despite these promising results, there are several analyses that the authors could incorporate to offer additional support to some of their claims (see weaknesses). In particular, this work would benefit from a more comprehensive comparison of the author's single recommender model vs an ensemble of specialist models, and the inclusion of 1-2 examples that showcase how their model could be translated into the clinic.

Strengths:

• A single AMR recommender system could potentially facilitate the adoption of MALDI-TOF-based antibiotic susceptibility profiling into clinical practices by reducing the number of models to be considered, and the efforts that may be required to periodically update them.

• Authors tested multiple combinations of embedders for the mass spectra and antibiotics while using different metrics to evaluate the performance of the resulting models. Models trained using different spectrum embedder-antibiotic embedder combinations had remarkably good performance for all tested metrics. The average ROC AUC scores for global and spectrum-level evaluations were above 0.9. Average ROC AUC scores for antibiotic-level evaluations were greater than 0.75.

• Authors showed that data collected in one location can be leveraged to improve the performance of models generated using a smaller number of samples collected at a different location. This result may encourage researchers to optimize data integration to reduce the burden of data generation for institutions interested in testing this method.

Weaknesses:

• Although ROC AUC is a widely used metric. Other metrics such as precision, recall, sensitivity, and specificity are not reported in this work. The last two metrics would help readers understand the model's potential implications in the context of clinical research.

• The authors did not hypothesize or describe in any way what an acceptable performance of their recommender system should be in order to be adopted by clinicians.

• Related to the previous comment, this work would strongly benefit from the inclusion of 1-2 real-life applications of their method that could showcase the benefits of their strategy for designing antibiotic treatment in a clinical setting.

• The authors do not offer information about the model features associated with resistance. This information may offer insights about mechanisms of antimicrobial resistance and how conserved they are across species.

• Comparison of AUC values across models lacks information regarding statistical significance. Without this information it is hard for a reader to figure out which differences are marginal and which ones are meaningful (for example, it is unclear if a difference in average AUC of 0.02 is significant). This applied to Figure 2, Figure 3, and Table 2 (and the associated supplementary figures).

• One key claim of this work was that their single recommender system outperformed specialist (single species-antibiotic) models. However, in its current status, it is not possible to determine that in fact that is the case (see comment above). Moreover, comparisons to species-level models (that combine all data and antibiotic susceptibility profiles for a given species) would help to illustrate the putative advantages of the dual branch neural network model over species-based models. This analysis will also inform the species (and perhaps datasets) for which specialist models would be useful to consider.

• Taking into account that the clustering of spectra embeddings seemed to be species-driven (Figure 4), one may hypothesize that there is limited transfer of information between species, and therefore the neural network model may be working as an ensemble of species models. Thus, this work would deeply benefit from a comparison between the authors' general model and an ensemble model in which the species is first identified and then the relevant species recommender is applied. If authors had identified cases to illustrate how data from one species positively influence the results for another species, they should include some of those examples.

Reviewer #2 (Public Review):

The authors frame the MS-spectrum-based prediction of antimicrobial resistance prediction as a drug recommendation task. Weis et al introduced the dataset this model is tested on and benchmark models which take as input a single species and are trained to predict resistance to a single drug. Instead here, a pair of drug and spectrum are fed to 2 neural network models to predict a resistance probability. In this manner, knowledge from different drugs and species can be shared through the model parameters. Three questions are asked: 1. what is the best way to encode the drugs? 2. does the dual NN outperform the single-spectrum drug?

Overall the paper is well-written and structured. It presents a novel framework for a relevant problem. The work would benefit from more work on evaluation.

Author response:

The following is the authors’ response to the previous reviews.

Reviewer 1:

• Although ROC AUC is a widely used metric. Other metrics such as precision, recall, sensitivity, and specificity are not reported in this work. The last two metrics would help readers understand the model’s potential implications in the context of clinical research.

In response to this comment and related ones by Reviewer 2, we have overhauled how we evaluate our models. In the revised version, we have removed Micro ROC-AUC, as this evaluation metric is hard to interpret in the recommender system setting. Instead, the updated version fully focuses on two metrics: ROC-AUC and Precision at 1 of the negative class, both computed per spectrum and then averaged (equivalent to the instance-wise metrics in the previous version of the manuscript). We believe these metrics best reflect the use-case of AMR recommenders. In addition, we have kept (drug-)macro ROC-AUC as a complementary evaluation metric. As the ROC-AUC can be decomposed into sensitivity and specificity (at different prediction probability thresholds), we have added a ROC curve where sensitivity and specificity are indicated in Figure 8 (Appendices).

• The authors did not hypothesize or describe in any way what an acceptable performance of their recommender system should be in order to be adopted by clinicians.

In Section 4.3, we have extended our experiments to include a baseline that represents a “simulated expert”. In short, given a species, an expert can already make some best guesses as to what drugs will be effective or not. To simulate this, we count resistance frequencies per species and per drug in the training set, and use this as predictions of a “simulated expert”.

We now mention in our manuscript that any performance above this level results in a real-world information gain for clinical diagnostic labs.

• Related to the previous comment, this work would strongly benefit from the inclusion of 1-2 real-life applications of their method that could showcase the benefits of their strategy for designing antibiotic treatment in a clinical setting.

While we think this would be valuable to try out, we are an in silico research lab, and the study we propose is an initial proof-of-concept focusing on the methodology. Because of this, we feel a real-life application of the model is out-of-scope for the present study.

• The authors do not offer information about the model features associated with resistance. This information may offer insights about mechanisms of antimicrobial resistance and how conserved they are across species.

In general, MALDI-TOF mass spectra are somewhat hard to interpret. Because of a limited body of work analyzing resistance mechanisms with MALDI-TOF MS, it is hard to link peaks back to specific pathways. For this reason, we have chosen to forego such an analysis. After all, as far as we know, typical MALDI-TOF MS manufacturers’ software for bacterial identification also does not provide interpretability results or insights into peaks, but merely gives an identification and confidence score.

However, we do feel that the whole topic revolving around “the degree of biological insight a data modality might give versus actual performance and usability” merits further discussion. We have ultimately decided not to include a segment in our discussion section as it is hard to discuss this matter concisely.

• Comparison of AUC values across models lacks information regarding statistical significance. Without this information it is hard for a reader to figure out which differences are marginal and which ones are meaningful (for example, it is unclear if a difference in average AUC of 0.02 is significant). This applied to Figure 2, Figure 3, and Table 2 (and the associated supplementary figures).

To make trends a bit more clear and easier to discern, in our revised manuscript, all models are run for 5 replicates (as opposed to 3 in the previous version).

There is an ongoing debate in the ML community whether statistical tests are useful for comparing machine learning models. A simple argument against them is that model runs are typically not independent from each other, as they are all trained on the same data. The assumptions of traditional statistical tests are therefore violated (t-test, Wilcoxon test, etc.). With such tests statistical significance of the smallest differences can simply be achieved by increasing the number of replicates (i.e. training the same models more times).

More complicated but more appropriate statistical tests also exist, such as the 5x2 cross-validated t-test of Dietterich: “Approximate statistical tests for comparing supervised classification learning algorithms”, Neural computation 1998. However, these tests are typically not considered in deep learning, because only 10% of the data can be used for training, which is practically not desirable. The Friedman test of Demšar "On the appropriateness of statistical tests in machine learning." Workshop on Evaluation Methods for Machine Learning in conjunction with ICML. 2008., in combination with posthoc pairwise tests, is still frequently used in machine learning, but that test is only applicable in studies where many datasets are tested.

For those reasons, most deep learning papers that only analyse a few datasets typically do not consider any statistical tests. For the same reasons, we are also not convinced of the added value of statistical tests in our study.

• One key claim of this work was that their single recommender system outperformed specialist (single species-antibiotic) models. However, in its current status, it is not possible to determine that in fact that is the case (see comment above). Moreover, comparisons to species-level models (that combine all data and antibiotic susceptibility profiles for a given species) would help to illustrate the putative advantages of the dual branch neural network model over species-based models. This analysis will also inform the species (and perhaps datasets) for which specialist models would be useful to consider.

We thank the reviewer for this excellent suggestion. In our new manuscript, we have dedicated an entire section of experiments to testing such species-specific recommender models (Section 4.2). We find that species-specific recommender systems generally outperform the models trained globally across all species. As a result, our manuscript has been majorly reworked.

• Taking into account that the clustering of spectra embeddings seemed to be species-driven (Figure 4), one may hypothesize that there is limited transfer of information between species, and therefore the neural network model may be working as an ensemble of species models. Thus, this work would deeply benefit from a comparison between the authors' general model and an ensemble model in which the species is first identified and then the relevant species recommender is applied. If authors had identified cases to illustrate how data from one species positively influence the results for another species, they should include some of those examples.

See the answer to the remark above.

• The authors should check that all abbreviations are properly introduced in the text so readers understand exactly what they mean. For example, the Prec@1 metric is a little confusing.

See the answer to a remark above for how we have overhauled our evaluation metrics in the revised version. In addition, in the revised version, we have bundled our explanations on evaluation metrics together in Section 3.2. We feel that having these explanations in a separate section will improve overall comprehensibility of the manuscript.

• The authors should include information about statistical significance in figures and tables that compare performance across models.

See answer above.

• An extra panel showing species labels would help readers understand Figure 11.

We have tried to play around with including species labels in these plots, but could not make it work without overcrowding the figure. Instead, we have added a reminder in the caption that readers should refer back to an earlier figure for species labels.

• The authors initially stated that molecular structure information is not informative. However, in a second analysis, the authors stated that molecular structures are useful for less common drugs. Please explain in more detail with specific examples what you mean.

In the previous version of our manuscript, we found that one-hot embedding-based models were superior to structure-based drug embedders for general performance. The latter however, delivered better transfer learning performance.

In our new experiments however, we perform early stopping on “spectrum-macro” ROC-AUC (as opposed to micro ROC-AUC in the previous version). As a consequence, our results are different. In the new version of our manuscript, Morgan Fingerprints-based drug embedders generally outperform others both “in general” and for transfer learning. Hence, our previously conflicting statements are not applicable to our new results.

• The authors may want to consider adding a few sentences that summarize the 'Related work' section into the introduction, and converting the 'Related work' section into an appendix.

While we acknowledge that such a section is uncommon in biology, in machine learning research, a “related work” section is very common. As this research lies on the intersection of the two, we have decided to keep the section as such.

Reviewer 2:

• Are the specialist models re-trained on the whole set of spectra? It was shown by Weis et al. that pooling spectra from different species hinders performance. It would then be better to compare directly to the models developed by Weis et al, using their splitting logic since it could be that the decay in performance from specialists comes from the pooling. See the section "Species-stratified learning yields superior predictions" in https://doi.org/10.1038/s41591-021-01619-9.

We train our “specialist” (or now-called “species-drug classifiers”) just as described in Weis et al.: All labels for a drug are taken, and then subsetted for a single species. We have clarified this a bit better in our new manuscript. The text now reads:

“Previous studies have studied AMR prediction in specific species-drug combinations. For this reason, it is useful to compare how the dual-branch setup weighs up against training separate models for separate species and drugs. In Weis et al. (2020b), for example, binary AMR classifiers are trained for the following three combinations: (1) E. coli with Ceftriaxone, (2) K. pneumoniae with Ceftriaxone, and (3) S. aureus with Oxacillin. Here, such "species-drug-specific classifiers" are trained for the 200 most-common combinations of species and drugs in the training dataset.

• Going back to Weis et al. a high variance in performance between species/drug pairs was observed. The metrics in Table 2 do not offer any measurement of variance or statistical testing. Indeed, some values are quite close e.g. Macro AUROC of Specialist MLP-XL vs One-hot M.

See our answer to a remark of Reviewer 1 for our viewpoint on statistical significance testing in machine learning.

• Since this is a recommendation task, why were no recommendation system metrics used, e.g. mAP@K, mRR, and so (apart from precision@1 for the negative class)? Additionally, since there is a high label imbalance in this task (~80% negatives) a simple model would achieve a very high precision@1.

See the answer to a remark above for how we have overhauled our evaluation metrics in the revised version. In addition, in choosing our metrics, we wanted metrics that are both (1) appropriate (i.e. recommender system metrics), but also (2) easy to interpret for clinicians. For this reason, we have not included metrics such as mAP@K or mRR. We feel that “spectrum-macro” ROC-AUC and precision@1 cover a sufficiently broad evaluation set of metrics but are easy enough to interpret.

• A highly similar approach was recently published (https://doi.org/10.1093/bioinformatics/btad717). Since it is quite close to the publication date of this paper, it could be discussed as concurrent work.

We thank the reviewer for bringing our attention to this study. We have added a paragraph in our revised version discussing this paper as concurrent work.

• It is difficult to observe a general trend from Figure 2. A statistical test would be advised here.

See our answer to a remark of Reviewer 1 for our viewpoint on statistical significance testing in machine learning.

• Figure 5. UMAPs generally don't lead to robust quantitative conclusions. However, the analysis of the embedding space is indeed interesting. Here I would recommend some quantitative measures directly using embedding distances to accompany the UMAP visualizations. E.g. clustering coefficients, distribution of pairwise distances, etc.

In accordance with this recommendation, we have computed many statistics on the MALDI-TOF spectra embedding spaces. However, we could not come up with any statistic that illuminated us more than the visualization itself. For this reason, we have kept this section as is, and let the figure speak for itself.

• Weis et al. also perform a transfer learning analysis. How does the transfer learning capacity of the proposed models differ from those in Weis et al?

Weis et al. perform experiments towards “transferability”, not actual transfer learning. In essence, they use a model trained on data from one diagnostic lab towards prediction on data from another. However, they do not conduct experiments to learn how much data such a pre-trained classifier needs to fine-tune it for adequate performance on the new diagnostic lab, as we do. The end of Section 4.4 discusses how our proposed models specifically shine in transfer learning. The paragraph reads:

“Lowering the amount of data required is paramount to expedite the uptake of AMR models in clinical diagnostics. The transfer learning qualities of dual-branch models may be ascribed to multiple properties. First of all, since different hospitals use much of the same drugs, transferred drug embedders allow for expressively representing drugs out of the box. Secondly, owing to multi-task learning, even with a limited number of spectra, a considerable fine-tuning dataset may be obtained, as all available data is "thrown on one pile".”

Version 2

Abstract

Abstract

Timely and effective use of antimicrobial drugs can improve patient outcomes, as well as help safeguard against resistance development. Matrix-assisted laser desorption/ionization time-of-flight mass spectrometry (MALDI-TOF MS) is currently routinely used in clinical diagnostics for rapid species identification. Mining additional data from said spectra in the form of antimicrobial resistance (AMR) profiles is, therefore, highly promising. Such AMR profiles could serve as a drop-in solution for drastically improving treatment efficiency, effectiveness, and costs.

This study endeavours to develop the first machine learning models capable of predicting AMR profiles for the whole repertoire of species and drugs encountered in clinical microbiology. The resulting models can be interpreted as drug recommender systems for infectious diseases. We find that our dual-branch method delivers considerably higher performance compared to previous approaches. In addition, experiments show that the models can be efficiently fine-tuned to data from other clinical laboratories. MALDI-TOF-based AMR recommender systems can, hence, greatly extend the value of MALDI-TOF MS for clinical diagnostics.

All code supporting this study is distributed on PyPI and is packaged under: https://github.com/gdewael/maldi-nn

Introduction

In diagnostic laboratories, matrix-assisted laser desorption/ionization time-of-flight mass spectrometry (MALDI-TOF MS) is routinely used for microbial species identification (Hou et al., 2019). Usually, microbial samples only require an overnight culturing step before being analyzed with mass spectrometry (Van Veen et al., 2010; Cuénod et al., 2021). Consequently, the technology provides a timeand costefficient way to accurately identify the pathogen underlying an infection.

Due to the rapid evolution of antibiotic resistant strains, it is increasingly difficult to determine a treatment based on only species identity. It has been estimated that infections caused by antibiotic-resistant bacteria have caused the deaths of 1.27 million people in 2019, making AMR one of the leading causes of death on earth (Murray et al., 2022). Projections have estimated that this annual number could rise to 10 million by 2050 (O’Neill, 2016), highlighting the need for responsible antimicrobial drug use. In light of this, diagnostic laboratories will often perform various tests, such as dilution arrays or disc diffusion tests, to probe which drug will be effective (Khan et al., 2019). Such experiments typically require further culturing and are either costly, labor-intensive, time-intensive, or a mixture of the above (Humphries, 2022).

Given that MALDI-TOF spectra are already routinely used for identification, it is worth investigating to which extent they can contain further information regarding the resistance status of strains (Weis et al., 2020a). Mining this information from the spectra could help inform healthcare workers of candidate drugs. This may nullify the need for phenotypical experiments, or (at least) direct the tests by narrowing down the choices. Furthermore, possessing a detailed resistance profile allows to treat with more specifically-working drugs (instead of broadspectrum antibiotics) (Weis et al., 2022). Consequently, predicting resistance status from MALDI-TOF spectra could help towards the goals of antibiotic stewardship (Shlaes et al., 1997).

It has been described that some known resistance mechanisms are outside of the m/z range that MALDI-TOF spectrometers can accurately measure (Humphries, 2022). Still, it remains largely unknown to which extent co-evolved traits, such as subtle changes in metabolism caused by the resistance mechanism, can be detected by MALDI-TOF spectra. A number of studies have shown that some resistant strains can reliably be predicted from MALDI-TOF MS, either by identifying and detecting specific markers (e.g. peaks) or by learning patterns from data (see §2.1). To our knowledge, all of these studies have modeled AMR prediction for specific species-drug combinations. For this reason, they learn very specific markers of resistance, not guaranteed to extrapolate well to other drugs and species. As susceptibility rapidly evolves, it is practically impossible to perform such studies for all clinically-relevant species-drug combinations. As such, the value of aforementioned studies remains of exploratory nature with limited practical value. In addition, their performance remains limited owing to small sample sizes and, likely, the inability of MALDI-TOF spectra to fully discriminate between the characteristics of interest (Bai et al., 2017). The recently published DRIAMS dataset (Weis et al., 2022) contains phenotypic AMR data covering a wide range of species and drugs, allowing to study MALDI-TOF-based AMR prediction on an unprecedented scale.

We posit that the most pertinent challenge healthcare workers face regarding AMR is to choose between all possible drugs given an infection, not whether one specific drug will be effective or not. For this reason, we argue that our models and evaluation metrics should be designed to optimally answer that question. In this study, a recommender model is proposed that can predict AMR for the whole range of pathogens and drugs encountered in clinical microbiology. In addition, species-specific recommender models for a range of common species are also trained. Our method jointly learns representations for antibiotic drugs and bacterial MALDITOF spectra. It can be used to recommend the mostlikely drug to work for any drug-spectrum combination. Consequently, the model is broadly-applicable and practical to use. To summarize, our contributions are as follows:

We formulate a dual-branch neural network recommender system for the prediction of AMR profiles. The model operates on MALDI-TOF spectra, as well as a representation of the candidate drug.

We evaluate multiple state-of-the-art techniques for representing drug identity in the model.

We compare “general” recommenders (trained on all spectra from all species) against speciesspecific recommender models

We perform evaluations by comparing our methods to non-recommender system baselines.

We show that the model efficiently transfers to data from diagnostic laboratories it wasn’t trained on. Making the model easy to adopt for hospitals lacking the means and/or volume to collect large data.

Related Work

MALDI-TOF-based machine learning

The most canonical task for MALDI-TOF-based methods is species identification. Identification solutions are usually provided by the MS manufacturers and are built on large, proprietary, in-house databases (Van Belkum et al., 2012). It is unclear how these closed-source identification pipelines work, but it is likely that query spectra are directly compared to the in-house database in an approach akin to nearest neighbors (Dauwalder et al., 2023). While this approach works excellently for identification of most species, some strains remain problematic (Cao et al., 2018; Vrioni et al., 2018). Furthermore, by presumably focusing on the presence or absence of specific peaks, a lot of spectral information stands unused (Florio et al., 2018).

For various difficult prediction cases, such as strain typing, researchers often resort to machine learning (Hettick et al., 2006; Wang et al., 2018; De Bruyne et al., 2011). Stifled by a historical lack of large open data, machine learning research on MALDI-TOF data remains in its infancy. Most studies have narrow scopes and simple datasets (e.g. binary classification), only warranting standard preprocessing and off-the-shelf learning techniques (Yu et al., 2022; Zhang et al., 2023; Chung et al., 2023). Only a handful of examples exist of more advanced learning techniques specifically adapted to a MALDI-TOF-based task (Mortier et al., 2021; Weis et al., 2020a; Vervier et al., 2015). For a more thorough overview of MALDI-TOF-based machine learning, readers are referred to the review of Weis et al. (2020b).

During peer-review, our attention was brought to a similar concurrent study by Visonà et al. (2023). Their study similarly shows that recommender systems-like models outperform more-narrowly trained singlespecies and single-drug models. Their analysis, however, remains limited to fingerprint-based molecular representations. In addition, in this work, we demonstrate transfer learning between hospitals.

Dual-branch neural networks

The idea of processing and combining two separate streams of information with two neural networks is applied in many fields of machine learning, collectively referred to as deep multi-target prediction (Waegeman et al., 2019; Iliadis et al., 2022).

In collaborative filtering, the goal is to predict the preference of a user to items (He et al., 2017). In its most elementary neural form, both users and items are represented by one-hot encodings, generating a model unable to make salient predictions for new users or items without having seen them during training. To solve this, a body of works exists on trying to communicate userand item-identity to the model via side-information encoded in features (Zheng et al., 2017).

Dual-branch neural networks are also prevalent in language and vision. Recent advances in (multimodal) contrastive learning of image (and text) representations often rely on two neural encoders to learn a matching score between two views of the same or discordant objects (Radford et al., 2021; Chen et al., 2020). Language retrieval systems typically compare input vectors with a database of key vectors, each derived from a neural network, using approximate nearest neighbor search techniques (Karpukhin et al., 2020). In biology, fields of research employing dualbranch neural networks include (1) drug-target interaction (Lee et al., 2019), (2) single-cell multi-omics analysis (Lance et al., 2022), and (3) transcription factor binding prediction (Yang et al., 2020), among countless others.

Most of these applications can, to varying extents, be interpreted as (collaborative filtering) recommender systems. For example, contrastive languageimage models have been used to retrieve the mostsemantically similar images to a piece of text (Beaumont, 2022).

Methods

Data

To train models, we use the recently published DRIAMS database, consisting of 765 048 AMR measurements derived from 55 773 spectra across four different hospitals, spanning in total 74 different drugs (Weis et al., 2022; These figures reflect the size of the dataset as downloaded from the original Dryad repository https://doi.org/10.5061/dryad.bzkh1899q, and after processing. For example, the number of spectra listed here corresponds to all spectra in DRIAMS for which there exists at least one AMR measurement. The total number of spectra in DRIAMS counts 250 070, but no labels are associated with these extra spectra. Further, the naming of drugs was further preprocessed such that every drug can be linked to a single chemical identifier. For more information on which drugs were merged and how this was performed, see Appendix A). Every drug is characterized by a canonical SMILES string obtained from PubChem (Kim et al., 2023). As in the original DRIAMS publication, AMR measurements are binarized according to the EUCAST norms per drug. Specifically, intermediate or resistant values are assigned a positive label, and susceptible samples a negative one. Furthermore, spectra are identically processed as in the original publication. Briefly, the following steps are performed: (1) square-root transformation of the intensities, (2) smoothing using a Savitzky-Golay filter with half-window size of 10, (3) baseline correction using 20 iterations of the SNIP algorithm, (4) trimming to the 2000-20000 Da range, (5) intensity calibration so that the total intensity sums to 1, and (6) binning the intensities by summing all values in intervals of 3 Da. After preprocessing, every spectrum is represented as a 6000-dimensional vector.

The main experiments concern models that are trained on data from one hospital only (DRIAMS-A, University Hospital Basel). All spectra and measurements derived from the other three hospitals in DRIAMS are left out for transfer learning experiments (see §4.4). Within DRIAMS-A, all spectra from before 2018 are allocated to the training set, and all spectra measured during 2018 are evenly split between validation and test set. This split in time reflects a realistic evaluation scenario, as models trained on historical data need to generalize to new patients possibly infected by newly-evolved strains. The final sizes of all splits are as follows: 409 395 labels across 28 331 spectra for the training set, 76 431 labels across 4 994 spectra for the validation set, and 76 133 labels across 4 999 spectra for the test set.

Metrics

The main objective of this study is to train models to effectively recommend treatments for patients. Hence, unless otherwise noted, metrics are computed on a per-patient basis, and then averaged. This is equivalent to macro-averaged metrics, but then computed per instance (spectrum), instead of per class (drug) (Waegeman et al., 2018). For simplicity, we omit the “macro” prefix from metrics, and unless otherwise indicated - always use spectrum-macro metrics.

The area under the receiver operating character- istic curve (ROC-AUC) measures the probability that any positive (resistant or intermediate) sample is assigned a higher predicted probability of being positive as compared to any negative (susceptible) sample. It is a measure of the average quality of the ranking of suggested drugs to a patient.

The Precision at 1 of the negative class (Prec@1(-)) evaluates how often the top-ranked prediction is correct. Hence, in this case, it reports the proportion of cases for which the “most-likely susceptible drug” prediction is actually an effective one. In a scenario where the top recommended drug is always administered, it corresponds to the percentage of correctlysuggested treatments.

Model architecture

We formulate AMR prediction as a multi-target classification problem with side-information for both instances and targets, also referred to as dyadic prediction (Waegeman et al., 2019). In this context, let us denote a sample in the dataset D by a triplet si, d j, yi j, where yi j denotes the resistance label of a microbial spectrum si ∈ {1,…,n} w.r.t. a drug d j ∈ {1,…,m}. This dataset can be arranged in an incomplete score matrixY ∈{0,1} n×m. In what follows, the final architectural set-ups used to present the results are described. For details on hyperparameter tuning, readers are referred to Appendix B.

The model consists of two separate neural network embedders Es (·) and Ed (·) for processing the spectra and drugs, respectively. The resulting instance and target embeddings xi and t j are then combined into a single score by their scaled dot product (Rendle et al., 2020). The scaling factor, with h the dimensionality of both embeddings, is inspired by the formulation of self-attention (Vaswani et al., 2017). It ensures the dot products to be of manageable magnitudes, even for large values of h. This score can be used together with the sigmoid function and the crossentropy loss to optimize the two-branch neural network to map a spectrum-drug pair to a resistance label (Iliadis et al., 2022). An overview of the model is visualized in Figure 1.

Architectural overview of the proposed model. AMR labels of spectrum-drug pairs can be represented in an incomplete matrix. A microbial sample that is susceptible to a drug is denoted by a negative label (orange), whereas positive labels (blue) signify an intermediate or resistant combination. Instance (spectrum) and target (drug) embeddings xi and t j are obtained from their respective input representations passed through their respective neural network branch. The two resulting embeddings are aggregated to a single score by their (scaled) dot product. The cross-entropy loss optimizes this score to be maximal or minimal for positive or negative combinations of microbial spectra and drugs, respectively.

The representations of the instance vectors xi are extracted from a neural network Es (·)operating on the processed and binned MALDI-TOF spectra si. Es (·) is parameterized by a multi-layer perceptron (MLP), consisting of a series of fully-connected layers. Between every two such layers, a series of operations consisting of (1) a GeLU activation (Hendrycks and Gimpel, 2016), (2) a dropout rate of 0.2 (Srivastava et al., 2014), and (3) layer normalization (Ba et al., 2016), is applied. We include multiple model sizes in our final results (Table 1). To make comparisons easier, all models output the same number of hidden dimensions that are used in the dot product, xi ∈ ℝ64.

Drug identity can be communicated to the model in a number of ways. In this work, we study the following different input representations d j and embedder Ed(·) combinations:

As indices in a one-hot encoding paired with a single linear layer.

As Extended Connectivity Fingerprints paired with a single linear layer.

As DeepSMILES strings (O’Boyle and Dalke, 2018) paired with a 1D convolutional neural network (CNN).

As DeepSMILES strings paired with a gated recurrent unit neural network (GRU).

As DeepSMILES strings paired with a transformer neural network.

As images paired with a 2D CNN.

As rows of a pre-computed string kernel on the SMILES strings (LINGO (Vidal et al., 2005)), paired with a single linear layer.

For all these combinations, the embedder outputs target embeddings t j ∈ ℝ64. For more details on the different drug embedders and their hyperparameters (as well as their tuning), see Appendix B. For every combination of spectrum embedder (four sizes: S, M, L, and XL) and drug embedder (seven types), six different learning rates ({ 1e-5, 5e-5, 1e-4, 5e-4, 1e-3, 5e-3 }) are tested. For all these different combinations, five models are trained (using different random seeds for model initialization and batching of data). For every spectrum and drug embedder combination, only results from the best learning rate are presented; that is, the learning rate resulting in the best average validation ROC-AUC for that combination.

All models are trained with the Adam optimizer (Kingma and Ba, 2014) for a maximum of 50 epochs with a batch size of 128. A linear learning rate warmup over the first 250 steps is applied, after which the rate is kept constant. As every epoch constitutes one pass over every label and, hence, multiple passes over every individual drug and spectrum, a branch can technically already be overfitting before the end of the first epoch. Because of this, performance on the validation set is checked every tenth of an epoch. Training is halted early when validation ROC-AUC hasn’t improved for 10 validation set checks. The checkpoint of the best performing model (in terms of validation ROC-AUC) is used as the final model.

Results

The following section will first relay the results of the different dual-branch model configurations. After, the “general” AMR recommender is matched up against “species-specific” and “species-drug-specific” models. Finally, the models’ capabilities and representations are examined through transfer learning and embeddings.

Encoding species and drugs effectively

Figure 2 shows the performance of all trained models in terms of their average ROC-AUC and Prec@1(-). It can be seen that, in general, performance differences between model configurations occupy a small margin. However, trends can still be found. Models using Morgan Fingerprints typically outperform other drug embedding strategies. Morgan Fingerprints provide a compressed and pre-processed input format, the nature of which provides an apparent advantage over input representations that require more pattern extraction. The small number of different antimicrobial drugs may not be conducive to learning complex representations. Indeed, embedding drugs without a compound information (i.e. one-hot embedding) is a competitive approach for this problem, resulting in the on average second best models in terms of ROCAUC. On the spectrum embedder side, it is observed that the medium or large variants typically perform best. The full ROC curve (showing sensitivity and specificity) for the best-performing model is shown in Figure 8.

Barplots showing test performance results for all trained models. ROC-AUC evaluates overall ranking of predictions. Prec@1(-) evaluates how often the top suggested treatment would be effective. Both metrics are calculated per spectrum/patient and then averaged. Errorbars represent the standard deviation over five random model seeds. The x-axis and colors show the different drug and spectrum embedders, respectively.

Performance in terms of Macro ROC-AUC can be found in Figure 9. The Macro ROC-AUC averages the ROC-AUC for every individual drug. Here, Morgan fingerprints similarly reach the best performances. The full list of performances can be found in Table 5.

In Figure 10, the performance of the spectrum embedder sizes is compared against a linear baseline. The linear baseline uses the same preprocessed input spectrum representation, but only uses a single linear combination to produce an embedding. For this comparison, only the Morgan fingerprint drug embedders are used, as they produce the best-performing models overall. Models using nonlinear multi-layer spectrum embedders obtain considerably better performance over linear embedders.

Species-specific models improve recommendation

The recommender systems presented in §4.1 provide an incredibly general tool. Trained as single models for all species and drugs, their versatility is unparalleled compared to previous studies that create classifiers for specific drug-species combinations (Weis et al., 2020b). In between the extremes of “one model for everything” and “a model per species and per drug”, there lies a compromising approach: a speciesspecific recommender system for all drugs. Such recommender systems would be more specialized in nature, but their usefulness hinges upon having done prior species identification. As these are typically included in the MS’ manufacturers software, a more specialized species-specific recommender may provide better performance without incurring extra cost. The disadvantage of such models is that (1) they can not be used for species for which there is not enough data to train a separate model (i.e. rarely occurring species), and (2) they rely on the prior identification step to be correct.

Here, we create species-specific recommender models for the 25 most-occurring species in DRIAMS-A. The training setup for these models is kept the same as in §4.1. The difference between “general” recommenders and “species-specific recommenders” is that each species-specific recommender model is only trained on the subset of data covering their respective species (As these models use a smaller training set, validation is checked every fourth of an epoch instead of every tenth). Together, the test predictions of the 25 species-specific recommenders cover 4229 spectra, 56 drugs, and 69827 AMR labels (covering 91.27% of the original test set). Table 2 compares the two best “general” recommenders from §4.1 to their species-specific recommender counterparts. It is observed that species-specific recommenders deliver better predictions across all evaluated metrics.

As opposed to the species-specific models, the “general” recommender can use learned representations from one species to enhance predictions for other species, benefitting from multi-task learning. The fact that this latter mode of learning performs worse on this problem, however, indicates that such transfer of learned knowledge is of limited usefulness for AMR prediction. Still, the “general” recommender model remains useful in instances where the species could not be identified, or is rare. In Table 6, the 25 species for which specific recommenders were trained are listed, along with their performances.

Dual-branch recommenders improve over baselines

In order to gain better insight into the performance of our models, in this section, both our “general” and “species-specific” recommenders are squared up against extensive baselines.

Previous studies have studied AMR prediction in specific species-drug combinations. For this reason, it is useful to compare how the dual-branch setup weighs up against training separate models for separate species and drugs. In Weis et al. (2020b), for example, binary AMR classifiers are trained for the following three combinations: (1) E. coli with Ceftriaxone, (2) K. pneumoniae with Ceftriaxone, and (3) S. aureus with Oxacillin. Here, such “species-drug-specific classifiers” are trained for the 200 most-common combinations of species and drugs in the training dataset. For these combinations, binary logistic regression, XGBoost (Chen and Guestrin, 2016) and MLPs are tested. The tested MLPs come in the same four sizes as the spectrum branches of the dual-branch models. Other than having an output node of size 1 for binary classification, they share all hyperparameters with the tested spectrum branches. For details on the training and tuning procedure of all baselines, see Appendix B.2.2.

There exist many species-drug combinations for which there are either only positive or only negative labels. As it is impossible to train and evaluate models for these cases, models are trained only for the 200 most-occurring combinations for which both labels are present in the training, validation and test set. We refer to these models as “species-drug classifiers”.

In addition, it is useful to probe model performance against what experts may be able to guess. Given knowledge of the species identity in question, an expert will in many cases already be able to make a good guess towards what drugs will be effective or not. Hence, baseline “best guess” performance would not result in a ROC-AUC of 0.5. A way to simulate such “expert’s best guess” baseline predictions is through counting label frequencies in the training set. More specifically, for a test label belonging to a certain species and drug, the labels in the training set corresponding to that drug and species can be gathered. The frequency by which that training set is positive or negative can be used to infer a test predicted probability. We refer to this baseline as “simulated expert’s best guess”.

Table 3 compares the recommenders from §4.2 to non-recommender baselines. As the baselines are only trained on the 200 most-common species-drug combinations, performance is computed on that subset of test labels. This reduced test set spans 4017 spectra, 35 drugs, and 53503 labels (covering 70.28% of the original test set). Dual-branch recommenders outperform baselines on all but one metric. Logistic regression baselines result in the best average ROCAUC for individual species-drug combinations. By all other metrics, dual-branch recommenders outshine a collection of species-drug-specific classifiers. It’s illustrated that, when the question is to choose between drugs for a patient (evaluated by the patient-averaged ROC-AUC or Prec@1(-)), a model designed as a recommender will outperform binary classification models trained to predict AMR for specific drugs. On the other hand, species-specific binary classifiers are optimal for distinguishing spectra for a specific drug. The crux of our case in favor of recommender models relies, hence, on the fact that patient-averaged metrics are more representative of AMR models’ utility in clinical diagnostics.

It is useful to note that any gain in performance over the “simulated expert” means that AMR signal could be mined from the spectra. Hence, any performance above this level results in a real-world information gain for clinical diagnostic laboratories.

Efficient transfer learning to new hospitals

An AMR prediction model trained using data from one hospital may not be suitable for use in other hospitals for several reasons. First, protocols such as sample preparation and culturing media differ from hospital to hospital, resulting in systematic differences in MALDI-TOF spectra (Weis et al., 2022). Second, epidemiology is spatially varied. Drug-resistant clades may be prevalent in one region or country, but absent in another (Humphries, 2022). Finally, the MALDITOF instruments themselves may also be specific to the hospital and influence the readout. This influences prediction models, as a hospital-specific effect is reported by the study introducing the DRIAMS dataset (Weis et al., 2022). They find that models typically perform best when trained with data from the same hospital. Here, hospital transferability is studied in the context of transfer learning.

Data from DRIAMS-B, -C and -D, are split into training, validation and test set. The train set for these hospitals consists of 1000 randomly-drawn spectra, simulating a small data scenario where the hospital has not spent considerable efforts in data collection. The remaining spectra for all three hospitals are evenly split among validation and test set.

For all three hospitals, we train models in the same way as previously (see §3.3). A comparison is made between fine-tuning starting from models trained on DRIAMS-A (i.e. models from previous sections) and dual-branch models trained from scratch (Figure 3). For simplicity, we transfer the non-species-specific, “general” recommenders, as we feel this reflects a more realistic use-case for labs that cannot afford to gather spectra for all possible species and additionally fine-tune them. Over all three hospitals, models finetuned from a DRIAMS-A checkpoint generally outperform models trained from scratch. This trend holds true over different numbers of spectra available in the fine-tuning set. In general, it can be seen that pretrained models require very little fine-tuning spectra to obtain performances in the same order of magnitude as with DRIAMS-A (§4.1). Performance comparisons of the same models in terms of other metrics are shown in Figure 11

Transfer learning of DRIAMS-A models to other hospitals. Errorbands show the standard deviation over five runs. Results in terms of other evaluation metrics are shown in Figure 11.

Lowering the amount of data required is paramount to expedite the uptake of AMR models in clinical diagnostics. The transfer learning qualities of dual-branch models may be ascribed to multiple properties. First of all, since different hospitals use much of the same drugs, transferred drug embedders allow for expressively representing drugs out of the box. Secondly, owing to multi-task learning, even with a limited number of spectra, a considerable fine-tuning dataset may be obtained, as all available data is “thrown on one pile”.

MALDI-TOF spectra embeddings

To investigate what the dual-branch models have learned to represent, MALDI-TOF spectra embeddings are examined. For this purpose, both the bestperforming “general” recommender and “speciesspecific” recommender are used. Here, we visualize the embeddings xi ∈ ℝ64 of all test set spectra from the 25 most-occurring pathogens. To visualize in a 2dimensional space, UMAP is applied (using default parameters apart from min_dist=0.5; increasing this parameter helps reduce UMAP packing points too tightly together, hence, making for a more-legible plot). Figure 4 shows the resulting embeddings, colored by species identity, as well as by their AMR status to a selection of drugs.

UMAP scatterplots of test set MALDI-TOF spectra embeddings xi. Top: Embeddings from a “general” (trained on all species) recommender. Only embeddings belonging to the 25 most-occurring species in the test set are shown. The panels on the right show the same embeddings as on the left, but colored according to its AMR status to a certain drug. The four displayed drugs are selected based on a ranking of the product of the number of positive and negative labels . In this way, the drugs that have a lot of observed labels, both positives and negatives, are displayed. Bottom: Highlighted embeddings from a S. epidermidis specific recommender model.

The MALDI-TOF embeddings from the “general” recommender model are grouped primarily per species. This shows that, without being instructed to discriminate between species, the model has learned to group spectra of the same species together. Furthermore, species under the same genus are typically grouped close together, illustrating that the model can pick up hierarchical relations in the tree of life from the data. Within species clusters, the AMR status subplots show that samples are often grouped according to their resistance. For example, for S. epidermidis and S. aureus, multidrug resistant variants clearly form subclusters. In addition, the cluster of E. coli spectra shows a clear tail with samples resistant to ciprofloxacin. Embeddings from the species-specific recommender models show this phenomenom more clearly. UMAP embedding plots from the “general recommender” colored by other drugs are shown in Figure 12. In addition, species-specific recommender system embeddings for some prominent species are shown in Figure 13.

Discussion

Prior work on AMR prediction has always modeled within the boundaries of one clade and drug(class), using standard machine learning practices. This work differentiates itself from others by constructing one model for the whole range of drugs encountered in clinical diagnostics. We propose to model AMR prediction via dual-branch neural networks, producing a novel MALDI-TOF-based AMR recommender system. The proposed models come with improved performance over the approaches taken in previous works.

In clinical diagnostics, AMR predictions could be used to decide which drug to administer on a perpatient basis. For this reason, we argue that evaluation metrics should probe the average quality of predictions per patient (i.e. spectrum-macro metrics). We show that, for these metrics, recommender systems consistently outperform baselines.

We postulate that the performance of the proposed models is still limited due to (1) lacking a MALDI-TOFspecific learning architecture, (2) collection of more data, especially on rarely-encountered species and drugs, and (3) inherent technological limitations of MALDI-TOF MS. Whilst the former is the subject of further machine learning research, the latter two can be considered by equipping the model with some notion of uncertainty, epistemic and aleatoric, respectively (Hüllermeier and Waegeman, 2021). In medical decision-making applications, effective uncertainty estimates would be an invaluable tool to aid understanding the models’ predictions. A fourth factor to consider is that perfect test set performance may also be unattainable due to labeling errors. This comes as a consequence of (1) error-prone laboratory measurements of MIC values, and (2) the fact that EUCAST norms change over time, resulting in outdated label thresholds for historical data.

As bacterial strains readily adapt resistance to new and frequently-used antibiotics, it is impossible for an AMR model to maintain its performance over time. Consequently, an obvious need for continual data collection and online machine learning approaches presents itself. It is for this reason that ML for AMR prediction will prove most valuable when integrated tightly in the inner workings of healthcare (Lee and Lee, 2020).

It stands to reason that blindly following the recommender system’s predictions spells misery. For example, healthcare practitioners should additionally take into account host-specific factors such as patient age, medical history, and concurrent medication. Additionally, as the model is trained on the whole repertoire of antimicrobial drugs, it will have learnt that broad-spectrum antibiotics are typically effective. Hence, it may overrecommend their use. As a consequence, the model’s proposed treatment strategies may not be aligned with antibiotic stewardship, instead exacerbating the very issue it is designed to mitigate. To tackle this problem, one could downweigh the prediction probabilities of undesirable drugs, or, alternatively, train a dual-branch model on only more-specifically-working drugs.

In summary, this study serves as the first proofof-concept for large MALDI-TOF-based antimicrobial drug recommenders. In this context, we highlight the need for appropriate metrics, proposing that perpatient metrics are most suitable. Extensive experiments on our proposed dual-branch model allow us to assemble some conclusions w.r.t. its use. Firstly, we find that medium-sized MLP spectrum embedders (counting 3.2M weights) generally perform best. Second, incorporating chemical information works best using Morgan Fingerprints. Third, while more data may skew the favor towards the other side, given the current available data, species-specific models outperform recommenders trained for all species. For the smaller datasets used in the transfer learning experiments, the structural inductive bias lent to the model via Morgan fingerprints delivers best results. Our experiments demonstrate that dual-branch recommenders outperform non-recommender baselines on relevant metrics. In the above discussion, some considerations are listed w.r.t. its practical implementation in healthcare. Taken together, this work demonstrates the potential of AMR recommenders to greatly extend the value of MALDI-TOF MS for clinical diagnostics.

eLife Assessment

This valuable study presents a machine learning model to recommend effective antimicrobial drugs from patients' samples analysed with mass spectrometry. The evidence supporting the claims of the authors is convincing, although including a measure of statistical significance to compare different proposed models would further strengthen the support. This work will be of interest to computational biologists, microbiologists, and clinicians.

Reviewer #1 (Public Review):

Summary:

De Waele et al. reported a dual-branch neural network model for predicting antibiotic resistance profiles using matrix-assisted laser desorption/ionization time-of-flight (MALDI-TOF) mass spectrometry data. Neural networks were trained on the recently available DRIAMS database of MALDI-TOF mass spectrometry data and their associated antibiotic susceptibility profiles. The authors used dual branch neural network to simultaneously represent information about mass spectra and antibiotics for a wide range of species and antibiotic combinations. The authors showed consistent performance of their strategy to predict antibiotic susceptibility for different spectrum and antibiotic representations (i.e., embedders). Remarkably, the authors showed how small datasets collected at one location can improve the performance of a model trained with limited data collected at a second location. The authors also showed that species-specific models (trained in multiple antibiotic resistance profiles) outperformed both the single recommender model and the individual species-antibiotic combination models. Despite the promising results, the authors should explain in more detail some of the analyses reported in the manuscript (see weaknesses).

Strengths:

• A single AMR recommender system could potentially facilitate the adoption of MALDI-TOF based antibiotic susceptibility profiling into clinical practices by reducing the number of models to be considered, and the efforts that may be required to periodically update them.

• Authors tested multiple combinations of embedders for the mass spectra and antibiotics while using different metrics to evaluate the performance of the resulting models. Models trained using different spectrum embedder-antibiotic embedder combinations had remarkably good performance for all tested metrics. The average ROC AUC scores for global and species-specific evaluations were above 0.8.

• Authors developed species-specific recommenders as an intermediate layer between the single recommender system and single species-antibiotic models. This intermediate approach achieved maximum performance (with one type of the species-specific recommender achieving a 0.9 ROC AUC), outlining the potential of this type of recommenders for frequent pathogens.

• Authors showed that data collected in one location can be leveraged to improve the performance of models generated using a smaller number of samples collected at a different location. This result may encourage researchers to optimize data integration to reduce the burden of data generation for institutions interested in testing this method.

Weaknesses:

• Section 4.3 ("expert baseline model"): the authors need to explain how the probabilities defined as baselines were exactly used to predict individual patient susceptible profiles.

• Authors do not offer information about the model features associated with resistance. Although I understand the difficulty of mapping mass spectra to specific pathways or metabolites, mechanistic insights are much more important in the context of AMR than in the context of bacterial identification. For example, this information may offer additional antimicrobial targets. Thus, authors should at least identify mass spectra peaks highly associated with resistance profiles. Are those peaks consistent across species? This would be a key step towards a proteomic survey of mechanisms of AMR. See previous work on this topic: PMIDs: 35586072 and 23297261.

Reviewer #2 (Public Review):

The authors frame the MS-spectrum-based prediction of antimicrobial resistance prediction as a drug recommendation task. Weis et al. introduced the dataset this model is tested on and benchmark models which take as input a single species and are trained to predict resistance to a single drug. Instead here, a pair of drugs and spectrum are fed to 2 neural network models to predict a resistance probability. In this manner, knowledge from different drugs and species can be shared through the model parameters. Questions asked: 1. what is the best way to encode the drugs? 2. does the dual NN outperform the single spectrum-drug?

Overall the paper is well-written and structured. It presents a novel framework for a relevant problem.

Author response:

The following is the authors’ response to the previous reviews.

Reviewer #1:

Section 4.3 ("expert baseline model"): the authors need to explain how the probabilities defined as baselines were exactly used to predict individual patient susceptible profiles.

We have added a more detailed and mathematically formal explanation of the “simulated expert’s best guess” in Section 4.3.

This section now reads:

“More formally, considering all training spectra as Strain, all training labels corresponding to one drug j and species t are gathered:

The "simulated expert's best guess" predicted probability for any spectrum si and drug dj, then, corresponds to, the fraction of positive labels in their corresponding training label set :

”

Authors should explain in more detail how a ROC curve is generated from a single spectrum (i.e., per patient) and then average across spectra. I have an idea of how it's done but I am not completely sure.

We have added a more detailed explanation in Section 3.2. It reads:

To compute the (per-patient average) ROC-AUC, for any spectrum/patient, all observed drug resistance labels and their corresponding predictions are gathered. Then, the patient-specific ROC-AUC is computed on that subset of labels and predictions. Finally, all ROC-AUCs per patient are averaged to a "spectrum-macro" ROC-AUC.

In addition, our description under Supplementary Figure 8 (showing the ROC curve) provides additional clarification:

Note that this ROC curve is not a traditional ROC curve constructed from one single label set and one corresponding prediction set. Rather, it is constructed from spectrum-macro metrics as follows: for any possible threshold value, binarize all predictions. Then, for every spectrum/patient independently, compute the sensitivity and specificity for the subset of labels corresponding to that spectrum/patient. Finally, those sensititivies and specificities are averaged across patients to obtain one point on above ROC curve.

Section 3.2 & reply # 1: can the authors compute and apply the Youden cutoff that gives max precision-sensitivity for each ROC curve? In that way the authors could report those values.

We have computed this cut-off on the curve shown in Supplementary Figure 8. The Figure now shows the sensitivity and specificity at the Youden cutoff in addition to the ROC. We have chosen only to report these values for this model as we did not want to inflate our manuscript with additional metrics (especially since the ROC-AUC already captures sensitivities and specificities). We do, however, see the value of adding this once, so that biologists have an indication of what kind of values to expect for these metrics.

Related to reply #5: assuming that different classifiers are trained in the same data, with the same number of replicates, could authors use the DeLong test compare ROC curves? If not, please explain why.

We thank the reviewer for bringing our attention to the DeLong’s test. It does indeed seem true that this test is appropriate for comparing two ROC-AUCs using the same ground truth values.

We have chosen not to use this test for one conceptual and one practical reason:

(1) Our point still stands that in machine learning one chooses the test set, and hence one can artificially increase statistical power by simply allocating a larger fraction of the data to test.

(2) DeLong’s test is defined for single AUCs (i.e. to compare two lists of predictions against one list of ground truths), but here we report the spectrum/patient-macro ROC-AUC. It is not clear how to adjust the test to macro-evaluated AUCs. One option may be to apply the test per patient ROC curve, and perform multiple testing correction, but then we are not comparing models, but models per patient. In addition, the number of labels/predictions per patient is prohibitively small for statistical power.

Reviewer #2 (Recommendations For The Authors):

After revision, all issues were been resolved.

Version 3

Abstract

Abstract

Timely and effective use of antimicrobial drugs can improve patient outcomes, as well as help safeguard against resistance development. Matrix-assisted laser desorption/ionization time-of-flight mass spectrometry (MALDI-TOF MS) is currently routinely used in clinical diagnostics for rapid species identification. Mining additional data from said spectra in the form of antimicrobial resistance (AMR) profiles is, therefore, highly promising. Such AMR profiles could serve as a drop-in solution for drastically improving treatment efficiency, effectiveness, and costs.

This study endeavours to develop the first machine learning models capable of predicting AMR profiles for the whole repertoire of species and drugs encountered in clinical microbiology. The resulting models can be interpreted as drug recommender systems for infectious diseases. We find that our dual-branch method delivers considerably higher performance compared to previous approaches. In addition, experiments show that the models can be efficiently fine-tuned to data from other clinical laboratories. MALDI-TOF-based AMR recommender systems can, hence, greatly extend the value of MALDI-TOF MS for clinical diagnostics.

All code supporting this study is distributed on PyPI and is packaged under: https://github.com/gdewael/maldi-nn

Introduction

In diagnostic laboratories, matrix-assisted laser desorption/ionization time-of-flight mass spectrometry (MALDI-TOF MS) is routinely used for microbial species identification (Hou et al., 2019). Usually, microbial samples only require an overnight culturing step before being analyzed with mass spectrometry (Van Veen et al., 2010; Cuénod et al., 2021). Consequently, the technology provides a time- and cost-efficient way to accurately identify the pathogen underlying an infection.

Due to the rapid evolution of antibiotic resistant strains, it is increasingly difficult to determine a treatment based on only species identity. It has been estimated that infections caused by antibiotic-resistant bacteria have caused the deaths of 1.27 million people in 2019, making AMR one of the leading causes of death on earth (Murray et al., 2022). Projections have estimated that this annual number could rise to 10 million by 2050 (O’Neill, 2016), highlighting the need for responsible antimicrobial drug use. In light of this, diagnostic laboratories will often perform various tests, such as dilution arrays or disc diffusion tests, to probe which drug will be effective (Khan et al., 2019). Such experiments typically require further culturing and are either costly, labor-intensive, time-intensive, or a mixture of the above (Humphries, 2022).

Given that MALDI-TOF spectra are already routinely used for identification, it is worth investigating to which extent they can contain further information regarding the resistance status of strains (Weis et al., 2020a). Mining this information from the spectra could help inform healthcare workers of candidate drugs. This may nullify the need for phenotypical experiments, or (at least) direct the tests by narrowing down the choices. Furthermore, possessing a detailed resistance profile allows to treat with more specifically-working drugs (instead of broad-spectrum antibiotics) (Weis et al., 2022). Consequently, predicting resistance status from MALDI-TOF spectra could help towards the goals of antibiotic stewardship (Shlaes et al., 1997).

It has been described that some known resistance mechanisms are outside of the m/z range that MALDI-TOF spectrometers can accurately measure (Humphries, 2022). Still, it remains largely unknown to which extent co-evolved traits, such as subtle changes in metabolism caused by the resistance mechanism, can be detected by MALDI-TOF spectra. A number of studies have shown that some resistant strains can reliably be predicted from MALDI-TOF MS, either by identifying and detecting specific markers (e.g. peaks) or by learning patterns from data (see §2.1). To our knowledge, all of these studies have modeled AMR prediction for specific species-drug combinations. For this reason, they learn very specific markers of resistance, not guaranteed to extrapolate well to other drugs and species. As susceptibility rapidly evolves, it is practically impossible to perform such studies for all clinically-relevant species-drug combinations. As such, the value of aforementioned studies remains of exploratory nature with limited practical value. In addition, their performance remains limited owing to small sample sizes and, likely, the inability of MALDI-TOF spectra to fully discriminate between the characteristics of interest (Bai et al., 2017). The recently published DRIAMS dataset (Weis et al., 2022) contains phenotypic AMR data covering a wide range of species and drugs, allowing to study MALDI-TOF-based AMR prediction on an unprecedented scale.

We posit that the most pertinent challenge healthcare workers face regarding AMR is to choose between all possible drugs given an infection, not whether one specific drug will be effective or not. For this reason, we argue that our models and evaluation metrics should be designed to optimally answer that question. In this study, a recommender model is proposed that can predict AMR for the whole range of pathogens and drugs encountered in clinical microbiology. In addition, species-specific recommender models for a range of common species are also trained. Our method jointly learns representations for antibiotic drugs and bacterial MALDI-TOF spectra. It can be used to recommend the mostlikely drug to work for any drug-spectrum combination. Consequently, the model is broadly-applicable and practical to use. To summarize, our contributions are as follows:

We formulate a dual-branch neural network recommender system for the prediction of AMR profiles. The model operates on MALDI-TOF spectra, as well as a representation of the candidate drug.

We evaluate multiple state-of-the-art techniques for representing drug identity in the model.

We compare “general” recommenders (trained on all spectra from all species) against speciesspecific recommender models

We perform evaluations by comparing our methods to non-recommender system baselines.

We show that the model efficiently transfers to data from diagnostic laboratories it wasn’t trained on. Making the model easy to adopt for hospitals lacking the means and/or volume to collect large data.

Related work

MALDI-TOF-based machine learning

The most canonical task for MALDI-TOF-based methods is species identification. Identification solutions are usually provided by the MS manufacturers and are built on large, proprietary, in-house databases (Van Belkum et al., 2012). It is unclear how these closed-source identification pipelines work, but it is likely that query spectra are directly compared to the in-house database in an approach akin to nearest neighbors (Dauwalder et al., 2023). While this approach works excellently for identification of most species, some strains remain problematic (Cao et al., 2018; Vrioni et al., 2018). Furthermore, by presumably focusing on the presence or absence of specific peaks, a lot of spectral information stands unused (Florio et al., 2018).

For various difficult prediction cases, such as strain typing, researchers often resort to machine learning (Hettick et al., 2006; Wang et al., 2018; De Bruyne et al., 2011). Stifled by a historical lack of large open data, machine learning research on MALDI-TOF data remains in its infancy. Most studies have narrow scopes and simple datasets (e.g. binary classification), only warranting standard preprocessing and off-the-shelf learning techniques (Yu et al., 2022; Zhang et al., 2023; Chung et al., 2023). Only a handful of examples exist of more advanced learning techniques specifically adapted to a MALDI-TOF-based task (Mortier et al., 2021; Weis et al., 2020a; Vervier et al., 2015). For a more thorough overview of MALDI-TOF-based machine learning, readers are referred to the review of Weis et al. (2020b).

During peer-review, our attention was brought to a similar concurrent study by Visonà et al. (2023). Their study similarly shows that recommender systems-like models outperform more-narrowly trained singlespecies and single-drug models. Their analysis, however, remains limited to fingerprint-based molecular representations. In addition, in this work, we demonstrate transfer learning between hospitals.

Dual-branch neural networks

The idea of processing and combining two separate streams of information with two neural networks is applied in many fields of machine learning, collectively referred to as deep multi-target prediction (Waegeman et al., 2019; Iliadis et al., 2022).

In collaborative filtering, the goal is to predict the preference of a user to items (He et al., 2017). In its most elementary neural form, both users and items are represented by one-hot encodings, generating a model unable to make salient predictions for new users or items without having seen them during training. To solve this, a body of works exists on trying to communicate user- and item-identity to the model via side-information encoded in features (Zheng et al., 2017).

Dual-branch neural networks are also prevalent in language and vision. Recent advances in (multimodal) contrastive learning of image (and text) representations often rely on two neural encoders to learn a matching score between two views of the same or discordant objects (Radford et al., 2021; Chen et al., 2020). Language retrieval systems typically compare input vectors with a database of key vectors, each derived from a neural network, using approximate nearest neighbor search techniques (Karpukhin et al., 2020). In biology, fields of research employing dualbranch neural networks include (1) drug-target interaction (Lee et al., 2019), (2) single-cell multi-omics analysis (Lance et al., 2022), and (3) transcription factor binding prediction (Yang et al., 2020), among countless others.

Most of these applications can, to varying extents, be interpreted as (collaborative filtering) recommender systems. For example, contrastive languageimage models have been used to retrieve the mostsemantically similar images to a piece of text (Beaumont, 2022).

Methods

Data

To train models, we use the recently published DRI-AMS database, consisting of 765 048 AMR measurements derived from 55 773 spectra across four different hospitals, spanning in total 74 different drugs (Weis et al., 2022; These figures reflect the size of the dataset as downloaded from the original Dryad repository https://doi.org/10.5061/dryad.bzkh1899q, and after processing. For example, the number of spectra listed here corresponds to all spectra in DRIAMS for which there exists at least one AMR measurement. The total number of spectra in DRIAMS counts 250 070, but no labels are associated with these extra spectra. Further, the naming of drugs was further preprocessed such that every drug can be linked to a single chemical identifier. For more information on which drugs were merged and how this was performed, see Appendix A). Every drug is characterized by a canonical SMILES string obtained from PubChem (Kim et al., 2023). As in the original DRIAMS publication, AMR measurements are binarized according to the EUCAST norms per drug. Specifically, intermediate or resistant values are assigned a positive label, and susceptible samples a negative one. Furthermore, spectra are identically processed as in the original publication. Briefly, the following steps are performed: (1) square-root transformation of the intensities, (2) smoothing using a Savitzky-Golay filter with half-window size of 10, (3) baseline correction using 20 iterations of the SNIP algorithm, (4) trimming to the 2000-20000 Da range, (5) intensity calibration so that the total intensity sums to 1, and (6) binning the intensities by summing all values in intervals of 3 Da. After preprocessing, every spectrum is represented as a 6000-dimensional vector.

The main experiments concern models that are trained on data from one hospital only (DRIAMS-A, University Hospital Basel). All spectra and measurements derived from the other three hospitals in DRI-AMS are left out for transfer learning experiments (see §4.4). Within DRIAMS-A, all spectra from before 2018 are allocated to the training set, and all spectra measured during 2018 are evenly split between validation and test set. This split in time reflects a realistic evaluation scenario, as models trained on historical data need to generalize to new patients possibly infected by newly-evolved strains. The final sizes of all splits are as follows: 409 395 labels across 28 331 spectra for the training set, 76 431 labels across 4 994 spectra for the validation set, and 76 133 labels across 4 999 spectra for the test set.

Metrics

The main objective of this study is to train models to effectively recommend treatments for patients. Hence, unless otherwise noted, metrics are computed on a per-patient basis, and then averaged. This is equivalent to macro-averaged metrics, but then computed per instance (spectrum), instead of per class (drug) (Waegeman et al., 2018). For simplicity, we omit the “macro” prefix from metrics, and - unless otherwise indicated - always use spectrum-macro metrics.

The area under the receiver operating characteristic curve (ROC-AUC) measures the probability that any positive (resistant or intermediate) sample is assigned a higher predicted probability of being positive as compared to any negative (susceptible) sample. It is a measure of the average quality of the ranking of suggested drugs to a patient. To compute the (perpatient average) ROC-AUC, for any spectrum/patient, all observed drug resistance labels and their corresponding predictions are gathered. Then, the patientspecific ROC-AUC is computed on that subset of labels and predictions. Finally, all ROC-AUCs per patient are averaged to a “spectrum-macro” ROC-AUC.

The Precision at 1 of the negative class (Prec@1(-)) evaluates how often the top-ranked prediction is correct. Hence, in this case, it reports the proportion of cases for which the “most-likely susceptible drug” prediction is actually an effective one. In a scenario where the top recommended drug is always administered, it corresponds to the percentage of correctly-suggested treatments.

Model architecture

We formulate AMR prediction as a multi-target classification problem with side-information for both instances and targets, also referred to as dyadic prediction (Waegeman et al., 2019). In this context, let us denote a sample in the dataset 𝒟 by a triplet (si, d j, yi j), where yi j denotes the resistance label of a microbial spectrum si ∈ { 1,…,n} w.r.t. a drug d j ∈ { 1,…,m}. This dataset can be arranged in an incomplete score matrixY ∈ {0, 1} n×m. In what follows, the final architectural set-ups used to present the results are described. For details on hyperparameter tuning, readers are referred to Appendix B.

The model consists of two separate neural network embedders Es (·) and Ed (·) for processing the spectra and drugs, respectively. The resulting instance and target embeddings xi and t j are then combined into a single score by their scaled dot product (Rendle et al., 2020). The scaling factor , h the with dimensionality of both embeddings, is inspired by the formulation of self-attention (Vaswani et al., 2017). It ensures the dot products to be of manageable magnitudes, even for large values of h. This score can be used together with the sigmoid function and the crossentropy loss to optimize the two-branch neural network to map a spectrum-drug pair to a resistance label (Iliadis et al., 2022). An overview of the model is visualized in Figure 1.

Architectural overview of the proposed model. AMR labels of spectrum-drug pairs can be represented in an incomplete matrix. A microbial sample that is susceptible to a drug is denoted by a negative label (orange), whereas positive labels (blue) signify an intermediate or resistant combination. Instance (spectrum) and target (drug) embeddings xi and t j are obtained from their respective input representations passed through their respective neural network branch. The two resulting embeddings are aggregated to a single score by their (scaled) dot product. The cross-entropy loss optimizes this score to be maximal or minimal for positive or negative combinations of microbial spectra and drugs, respectively.

The representations of the instance vectors xi are extracted from a neural network Es (·) operating on the processed and binned MALDI-TOF spectra si.

Es (·) is parameterized by a multi-layer perceptron (MLP), consisting of a series of fully-connected layers. Between every two such layers, a series of operations consisting of (1) a GeLU activation (Hendrycks and Gimpel, 2016), (2) a dropout rate of 0.2 (Srivastava et al., 2014), and (3) layer normalization (Ba et al., 2016), is applied. We include multiple model sizes in our final results (Table 1). To make comparisons easier, all models output the same number of hidden dimensions that are used in the dot product, xi ∈ ℝ64.

Drug identity can be communicated to the model in a number of ways. In this work, we study the following different input representations d j and embedder Ed (·) combinations:

As indices in a one-hot encoding paired with a single linear layer.

As Extended Connectivity Fingerprints paired with a single linear layer.

As DeepSMILES strings (O’Boyle and Dalke, 2018) paired with a 1D convolutional neural network (CNN).

As DeepSMILES strings paired with a gated recurrent unit neural network (GRU).

As DeepSMILES strings paired with a transformer neural network.

As images paired with a 2D CNN.

As rows of a pre-computed string kernel on the SMILES strings (LINGO (Vidal et al., 2005)), paired with a single linear layer.

For all these combinations, the embedder outputs target embeddings t j ∈ ℝ64. For more details on the different drug embedders and their hyperparameters (as well as their tuning), see Appendix B. For every combination of spectrum embedder (four sizes: S, M, L, and XL) and drug embedder (seven types), six different learning rates ({1e-5, 5e-5, 1e-4, 5e-4, 1e-3, 5e-3}) are tested. For all these different combinations, five models are trained (using different random seeds for model initialization and batching of data). For every spectrum and drug embedder combination, only results from the best learning rate are presented; that is, the learning rate resulting in the best average validation ROC-AUC for that combination.

All models are trained with the Adam optimizer (Kingma and Ba, 2014) for a maximum of 50 epochs with a batch size of 128. A linear learning rate warmup over the first 250 steps is applied, after which the rate is kept constant. As every epoch constitutes one pass over every label and, hence, multiple passes over every individual drug and spectrum, a branch can technically already be overfitting before the end of the first epoch. Because of this, performance on the validation set is checked every tenth of an epoch. Training is halted early when validation ROC-AUC hasn’t improved for 10 validation set checks. The checkpoint of the best performing model (in terms of validation ROC-AUC) is used as the final model.

Results

The following section will first relay the results of the different dual-branch model configurations. After, the “general” AMR recommender is matched up against “species-specific” and “species-drug-specific” models. Finally, the models’ capabilities and representations are examined through transfer learning and embeddings.

Encoding species and drugs effectively

Figure 2 shows the performance of all trained models in terms of their average ROC-AUC and Prec@1(-). It can be seen that, in general, performance differences between model configurations occupy a small margin. However, trends can still be found. Models using Morgan Fingerprints typically outperform other drug embedding strategies. Morgan Fingerprints provide a compressed and pre-processed input format, the nature of which provides an apparent advantage over input representations that require more pattern extraction. The small number of different antimicrobial drugs may not be conducive to learning complex representations. Indeed, embedding drugs without a compound information (i.e. one-hot embedding) is a competitive approach for this problem, resulting in the - on average - second best models in terms of ROC-AUC. On the spectrum embedder side, it is observed that the medium or large variants typically perform best. The full ROC curve (showing sensitivity and specificity) for the best-performing model is shown in Appendix C Figure 8.

Barplots showing test performance results for all trained models. ROC-AUC evaluates overall ranking of predictions. Prec@1(-) evaluates how often the top suggested treatment would be effective. Both metrics are calculated per spectrum/patient and then averaged. Errorbars represent the standard deviation over five random model seeds. The x-axis and colors show the different drug and spectrum embedders, respectively.

Performance in terms of Macro ROC-AUC can be found in Appendix C Figure 9. The Macro ROC-AUC averages the ROC-AUC for every individual drug. Here, Morgan fingerprints similarly reach the best performances. The full list of performances can be found in Appendix C Table 5.

In Appendix C Figure 10, the performance of the spectrum embedder sizes is compared against a linear baseline. The linear baseline uses the same pre-processed input spectrum representation, but only uses a single linear combination to produce an embedding. For this comparison, only the Morgan fingerprint drug embedders are used, as they produce the best-performing models overall. Models using non-linear multi-layer spectrum embedders obtain considerably better performance over linear embedders.

Species-specific models improve recommendation

The recommender systems presented in §4.1 provide an incredibly general tool. Trained as single models for all species and drugs, their versatility is unparalleled compared to previous studies that create classifiers for specific drug-species combinations (Weis et al., 2020b). In between the extremes of “one model for everything” and “a model per species and per drug”, there lies a compromising approach: a speciesspecific recommender system for all drugs. Such recommender systems would be more specialized in nature, but their usefulness hinges upon having done prior species identification. As these are typically included in the MS’ manufacturers software, a more specialized species-specific recommender may provide better performance without incurring extra cost. The disadvantage of such models is that (1) they can not be used for species for which there is not enough data to train a separate model (i.e. rarely occurring species), and (2) they rely on the prior identification step to be correct.

Here, we create species-specific recommender models for the 25 most-occurring species in DRIAMS-A. The training setup for these models is kept the same as in §4.1. The difference between “general” recommenders and “species-specific recommenders” is that each species-specific recommender model is only trained on the subset of data covering their respective species (As these models use a smaller training set, validation is checked every fourth of an epoch instead of every tenth.). Together, the test predictions of the 25 species-specific recommenders cover 4229 spectra, 56 drugs, and 69827 AMR labels (covering 91.27% of the original test set). Table 2 compares the two best “general” recommenders from §4.1 to their species-specific recommender counterparts. It is observed that species-specific recommenders deliver better predictions across all evaluated metrics.

As opposed to the species-specific models, the “general” recommender can use learned representations from one species to enhance predictions for other species, benefitting from multi-task learning. The fact that this latter mode of learning performs worse on this problem, however, indicates that such transfer of learned knowledge is of limited usefulness for AMR prediction. Still, the “general” recommender model remains useful in instances where the species could not be identified, or is rare. In Appendix C Table 6, the 25 species for which specific recommenders were trained are listed, along with their performances.

Dual-branch recommenders improve over baselines

In order to gain better insight into the performance of our models, in this section, both our “general” and “species-specific” recommenders are squared up against extensive baselines.

Previous studies have studied AMR prediction in specific species-drug combinations. For this reason, it is useful to compare how the dual-branch setup weighs up against training separate models for separate species and drugs. In Weis et al. (2020b), for example, binary AMR classifiers are trained for the following three combinations: (1) E. coli with Ceftriaxone, (2) K. pneumoniae with Ceftriaxone, and (3) S. aureus with Oxacillin. Here, such “species-drug-specific classifiers” are trained for the 200 most-common combinations of species and drugs in the training dataset. For these combinations, binary logistic regression, XGBoost (Chen and Guestrin, 2016) and MLPs are tested. The tested MLPs come in the same four sizes as the spectrum branches of the dual-branch models. Other than having an output node of size 1 for binary classification, they share all hyperparameters with the tested spectrum branches. For details on the training and tuning procedure of all baselines, see Appendix B.2.2.

There exist many species-drug combinations for which there are either only positive or only negative labels. As it is impossible to train and evaluate models for these cases, models are trained only for the 200 most-occurring combinations for which both labels are present in the training, validation and test set. We refer to these models as “species-drug classifiers”.

In addition, it is useful to probe model performance against what experts may be able to guess. Given knowledge of the species identity in question, an expert will - in many cases - already be able to make a good guess towards what drugs will be effective or not. Hence, baseline “best guess” performance would not result in a ROC-AUC of 0.5. A way to simulate such “expert’s best guess” baseline predictions is through counting label frequencies in the training set. More specifically, for a test label belonging to a certain species and drug, the labels in the training set corresponding to that drug and species can be gathered. The frequency by which that training set is positive or negative can be used to infer a test predicted probability. We refer to this baseline as “simulated expert’s best guess”. More formally, considering all training spectra as Strain, all training labels corresponding to one drug j and species t are gathered: The “simulated expert’s best guess” predicted probability for any spectrum si and drug d j, then, corresponds to, the fraction of positive labels in their corresponding training label set: Table 3 compares the recommenders from §4.2 to non-recommender baselines. As the baselines are only trained on the 200 most-common species-drug combinations, performance is computed on that subset of test labels. This reduced test set spans 4017 spectra, 35 drugs, and 53503 labels (covering 70.28% of the original test set). Dual-branch recommenders outperform baselines on all but one metric. Logistic regression baselines result in the best average ROC-AUC for individual species-drug combinations. By all other metrics, dual-branch recommenders outshine a collection of species-drug-specific classifiers. It’s illustrated that, when the question is to choose between drugs for a patient (evaluated by the patient-averaged ROC-AUC or Prec@1(-)), a model designed as a recommender will outperform binary classification models trained to predict AMR for specific drugs. On the other hand, species-specific binary classifiers are optimal for distinguishing spectra for a specific drug. The crux of our case in favor of recommender models relies, hence, on the fact that patient-averaged metrics are more representative of AMR models’ utility in clinical diagnostics.

It is useful to note that any gain in performance over the “simulated expert” means that AMR signal could be mined from the spectra. Hence, any performance above this level results in a real-world information gain for clinical diagnostic laboratories.

Efficient transfer learning to new hospitals

An AMR prediction model trained using data from one hospital may not be suitable for use in other hospitals for several reasons. First, protocols such as sample preparation and culturing media differ from hospital to hospital, resulting in systematic differences in MALDI-TOF spectra (Weis et al., 2022). Second, epidemiology is spatially varied. Drug-resistant clades may be prevalent in one region or country, but absent in another (Humphries, 2022). Finally, the MALDI-TOF instruments themselves may also be specific to the hospital and influence the readout. This influences prediction models, as a hospital-specific effect is reported by the study introducing the DRIAMS dataset (Weis et al., 2022). They find that models typically perform best when trained with data from the same hospital. Here, hospital transferability is studied in the context of transfer learning.

Data from DRIAMS-B, -C and -D, are split into training, validation and test set. The train set for these hospitals consists of 1000 randomly-drawn spectra, simulating a small data scenario where the hospital has not spent considerable efforts in data collection. The remaining spectra for all three hospitals are evenly split among validation and test set.

For all three hospitals, we train models in the same way as previously (see §3.3). A comparison is made between fine-tuning starting from models trained on DRIAMS-A (i.e. models from previous sections) and dual-branch models trained from scratch (Figure 3). For simplicity, we transfer the non-species-specific, “general” recommenders, as we feel this reflects a more realistic use-case for labs that cannot afford to gather spectra for all possible species and additionally fine-tune them. Over all three hospitals, models finetuned from a DRIAMS-A checkpoint generally outperform models trained from scratch. This trend holds true over different numbers of spectra available in the fine-tuning set. In general, it can be seen that pretrained models require very little fine-tuning spectra to obtain performances in the same order of magnitude as with DRIAMS-A (§4.1). Performance comparisons of the same models in terms of other metrics are shown in Appendix C Figure 11

Transfer learning of DRIAMS-A models to other hospitals. Errorbands show the standard deviation over five runs. Results in terms of other evaluation metrics are shown in Appendix C Figure 11.

Lowering the amount of data required is paramount to expedite the uptake of AMR models in clinical diagnostics. The transfer learning qualities of dual-branch models may be ascribed to multiple properties. First of all, since different hospitals use much of the same drugs, transferred drug embedders allow for expressively representing drugs out of the box. Secondly, owing to multi-task learning, even with a limited number of spectra, a considerable fine-tuning dataset may be obtained, as all available data is “thrown on one pile”.

MALDI-TOF spectra embeddings

To investigate what the dual-branch models have learned to represent, MALDI-TOF spectra embeddings are examined. For this purpose, both the bestperforming “general” recommender and “speciesspecific” recommender are used. Here, we visualize the embeddings xi ∈ ℝ64 of all test set spectra from the 25 most-occurring pathogens. To visualize in a 2-dimensional space, UMAP is applied (using default parameters apart from min_dist=0.5; Increasing this parameter helps reduce UMAP packing points too tightly together, hence, making for a more-legible plot.). Figure 4 shows the resulting embeddings, colored by species identity, as well as by their AMR status to a selection of drugs.

UMAP scatterplots of test set MALDI-TOF spectra embeddings xi. Top: Embeddings from a “general” (trained on all species) recommender. Only embeddings belonging to the 25 most-occurring species in the test set are shown. The panels on the right show the same embeddings as on the left, but colored according to its AMR status to a certain drug. The four displayed drugs are selected based on a ranking of the product of the number of positive and negative labels .In this way, the drugs that have a lot of observed labels, both positives and negatives, are displayed. Bottom: Highlighted embeddings from a S. epidermidis specific recommender model.

The MALDI-TOF embeddings from the “general” recommender model are grouped primarily per species. This shows that, without being instructed to discriminate between species, the model has learned to group spectra of the same species together. Furthermore, species under the same genus are typically grouped close together, illustrating that the model can pick up hierarchical relations in the tree of life from the data. Within species clusters, the AMR status subplots show that samples are often grouped according to their resistance. For example, for S. epidermidis and S. aureus, multidrug resistant variants clearly form subclusters. In addition, the cluster of E. coli spectra shows a clear tail with samples resistant to ciprofloxacin. Embeddings from the species-specific recommender models show this phenomenom more clearly. UMAP embedding plots from the “general recommender” colored by other drugs are shown in Appendix C Figure 12. In addition, species-specific recommender system embeddings for some prominent species are shown in Appendix C Figure 13.

Discussion

Prior work on AMR prediction has always modeled within the boundaries of one clade and drug(class), using standard machine learning practices. This work differentiates itself from others by constructing one model for the whole range of drugs encountered in clinical diagnostics. We propose to model AMR prediction via dual-branch neural networks, producing a novel MALDI-TOF-based AMR recommender system. The proposed models come with improved performance over the approaches taken in previous works.

In clinical diagnostics, AMR predictions could be used to decide which drug to administer on a perpatient basis. For this reason, we argue that evaluation metrics should probe the average quality of predictions per patient (i.e. spectrum-macro metrics). We show that, for these metrics, recommender systems consistently outperform baselines.

We postulate that the performance of the proposed models is still limited due to (1) lacking a MALDI-TOF-specific learning architecture, (2) collection of more data, especially on rarely-encountered species and drugs, and (3) inherent technological limitations of MALDI-TOF MS. Whilst the former is the subject of further machine learning research, the latter two can be considered by equipping the model with some notion of uncertainty, epistemic and aleatoric, respectively (Hüllermeier and Waegeman, 2021). In medical decision-making applications, effective uncertainty estimates would be an invaluable tool to aid understanding the models’ predictions. A fourth factor to consider is that perfect test set performance may also be unattainable due to labeling errors. This comes as a consequence of (1) error-prone laboratory measurements of MIC values, and (2) the fact that EUCAST norms change over time, resulting in outdated label thresholds for historical data.

As bacterial strains readily adapt resistance to new and frequently-used antibiotics, it is impossible for an AMR model to maintain its performance over time. Consequently, an obvious need for continual data collection and online machine learning approaches presents itself. It is for this reason that ML for AMR prediction will prove most valuable when integrated tightly in the inner workings of healthcare (Lee and Lee, 2020).

It stands to reason that blindly following the recommender system’s predictions spells misery. For example, healthcare practitioners should additionally take into account host-specific factors such as patient age, medical history, and concurrent medication. Additionally, as the model is trained on the whole repertoire of antimicrobial drugs, it will have learnt that broad-spectrum antibiotics are typically effective. Hence, it may overrecommend their use. As a consequence, the model’s proposed treatment strategies may not be aligned with antibiotic stewardship, instead exacerbating the very issue it is designed to mitigate. To tackle this problem, one could downweigh the prediction probabilities of undesirable drugs, or, alternatively, train a dual-branch model on only more-specifically-working drugs.

In summary, this study serves as the first proof-of-concept for large MALDI-TOF-based antimicrobial drug recommenders. In this context, we highlight the need for appropriate metrics, proposing that perpatient metrics are most suitable. Extensive experiments on our proposed dual-branch model allow us to assemble some conclusions w.r.t. its use. Firstly, we find that medium-sized MLP spectrum embedders (counting 3.2M weights) generally perform best. Second, incorporating chemical information works best using Morgan Fingerprints. Third, while more data may skew the favor towards the other side, given the current available data, species-specific models outperform recommenders trained for all species. For the smaller datasets used in the transfer learning experiments, the structural inductive bias lent to the model via Morgan fingerprints delivers best results. Our experiments demonstrate that dual-branch recommenders outperform non-recommender baselines on relevant metrics. In the above discussion, some considerations are listed w.r.t. its practical implementation in healthcare. Taken together, this work demonstrates the potential of AMR recommenders to greatly extend the value of MALDI-TOF MS for clinical diagnostics.